\documentclass{clv2025}

\jvol{vv}
\jnum{nn}
\jyear{2026}
\jinfo{}
\renewcommand{\footmark}{}
\dochead{Position Paper}

\usepackage[T1]{fontenc}
\usepackage[american]{babel}
\usepackage{natbib}
\bibliographystyle{compling}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}
\hypersetup{hidelinks}

\newcommand{\TC}{\textsc{TC}}
\newcommand{\LD}{\textsc{LD}}
\newcommand{\Hall}{\textsc{H}}
\newcommand{\SUP}{\textsc{SUP}}
\newcommand{\SV}{\textsc{SV}}
\newcolumntype{Y}{>{\raggedright\arraybackslash}X}
\newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}

\definecolor{clrH}{RGB}{155,18,18}
\definecolor{clrLD}{RGB}{22,98,48}
\definecolor{clrB}{RGB}{22,52,125}
\definecolor{clrP}{RGB}{92,61,145}
\definecolor{clrO}{RGB}{170,91,24}
\definecolor{clrG}{RGB}{65,65,65}
\definecolor{bgB}{RGB}{245,247,254}
\tikzset{
  input/.style={rectangle,rounded corners=4pt,draw=gray!55,fill=gray!8,
    font=\small\itshape,text width=5.0cm,align=center,inner sep=6pt},
  decision/.style={rectangle,rounded corners=3pt,draw=clrB!55,fill=bgB,
    font=\small,text width=5.2cm,align=center,inner sep=6pt},
  verdict/.style={rectangle,rounded corners=3pt,draw=clrG,fill=gray!8,
    font=\small\bfseries\sffamily,inner xsep=7pt,inner ysep=4pt},
  verdictH/.style={verdict,draw=clrH,text=clrH,fill=clrH!7},
  verdictLD/.style={verdict,draw=clrLD,text=clrLD,fill=clrLD!7},
  arrow/.style={-{Stealth[length=5pt,width=3.5pt]},thick,color=gray!70!black},
  edge/.style={font=\scriptsize\sffamily,fill=white,inner sep=1.5pt}
}

\title{Hallucination Is Contract-Relative: A Truth-Contract Framework for LLM Evaluation}
\runningtitle{Hallucination Is Contract-Relative}
\runningauthor{Carlesso et al.}

\author{Mathis Carlesso$^{1,2}$, Martino Lovisetto$^{2}$, El\H{o}d Egyed-Zsigmond$^{1}$, Pierre-Yves Genest$^{2}$, Stefan Duffner$^{1}$}

\affilblock{
  \affil{INSA Lyon, CNRS, LIRIS, UMR5205, Universit\'e Claude Bernard Lyon 1, 69621 Villeurbanne, France\\
  \quad \email{mathis.carlesso@insa-lyon.fr}}
  \affil{Alteca, 69100 Villeurbanne, France}
}

\begin{document}
\maketitle

\begin{abstract}
Hallucination in large language models is usually defined as unsupported or contradicted content.
This definition is useful for factual reporting but ambiguous when a task allows invention or requests figurative language.
We argue that hallucination should instead be a claim-level label relative to a task's \emph{truth contract}, $\TC(p)=(O_p,\sigma_p,K_p)$.
The contract specifies reference evidence $O_p$, a requested style level $\sigma_p$, and a content-permission policy $K_p$.
Its permission level $\kappa_p\in\{0,1,2\}$ distinguishes factual-only tasks, limited divergence, and in-frame invention, while further components specify where and how divergence is allowed.
The evaluator first separates factual claims from stylistic wording and then checks only the extracted claims.
The requested style level guides the interpretation of metaphors and other marked forms, but never permits unsupported content by itself.
A supported claim receives \emph{supported} (\SUP); a contradicted or unlicensed claim receives \emph{hallucination} (\Hall); and an evidence-unknown but properly licensed claim receives \emph{licensed divergence} (\LD).
These claim-level labels do not by themselves score a complete response, and no aggregation rule is known to be optimal across truth contracts.
We support the position with an author-coded mapping of forty evaluation resources and five worked cases.
In this sample, factuality benchmarks cluster around neutral language ($\sigma=0$) and factual-only permission ($\kappa=0$), whereas creative-writing benchmarks cluster around marked language ($\sigma=1$ or $2$) and task-implied invention ($\kappa=2$).
The two literatures therefore evaluate different truth contracts rather than conflicting versions of the same task.
The missing test is a controlled benchmark that varies style and permission independently.
\end{abstract}

\section{Introduction}\label{sec:intro}

Large language models (LLMs) can produce fluent claims that are unsupported by evidence or contradicted by known facts \citep{ji_survey_2023,huang_survey_2025}.
These errors limit the use of LLMs in high-stakes settings, and a large body of work aims to detect and reduce them \citep{lin_truthfulqa_2022,li_halueval_2023,wei_measuring_2024}.

However, unsupported generation is not always an error.
Fiction, brainstorming, and hypothesis generation require the model to go beyond established facts or provided documents \citep{franceschelli_creativity_2024,jiang_survey_2024,sui_confabulation_2024}.
A medical assistant should not invent a diagnosis.
A fiction-writing assistant must be able to invent characters and events.
The relevant distinction is therefore whether the task permits that departure from support, not merely whether a statement is supported.

Many current evaluation protocols do not represent this distinction explicitly.
Factuality and faithfulness evaluations typically penalize unsupported claims under a fixed evidence source \citep{maynez_faithfulness_2020,bang_hallulens_2025}.
Creativity evaluations reward novelty and appropriateness, but they do not always test whether an apparently novel statement violates a factual constraint \citep{acar_creativity_2019,hou_creativityprism_2025}.
Consequently, two protocols can assign different claim labels to the same wording because each silently assumes a different task.

A second diagnostic problem concerns form rather than invention.
Stylistic foregrounding makes linguistic form salient through devices such as deviation, parallelism, metaphor, and unusual phrasing \citep{leech_short_style_1981,vanpeer_stylistics_2020}.
Some of these devices can change how a claim is expressed without adding a new factual commitment \citep{badathala_match_2023,jin_deep_2022}.
An evaluator that literalizes a metaphor, or reads surface unexpectedness as evidence of factual deviation, can therefore report a hallucination although the underlying claim remains supported.
The point is not that all current factuality systems make this error, but that many protocols do not expose the requested style level as a variable, so the error cannot be measured.

\paragraph{Our position}
Hallucination is not a context-free property of a sentence.
It is a claim-level label relative to the task's truth contract.
We write this contract as $\TC(p)=(O_p,\sigma_p,K_p)$.
The reference evidence $O_p$ specifies the standard of support.
The requested style level $\sigma_p$ specifies how strongly the prompt asks for departure from neutral wording, with factual commitments preserved.
The content-permission policy $K_p$ specifies the type, scope, and required epistemic presentation of any permitted departure.
An unsupported claim is \Hall\ if it violates $K_p$; an evidence-unknown claim is \LD\ if $K_p$ permits it and its presentation satisfies the contract.
These labels apply to individual claims; assessing a complete response requires a separate aggregation step that this paper does not fix.
The contract does not change what is true.
It determines whether the model's departure from the relevant evidence violates the task.

Figure~\ref{fig:flip} illustrates this position.
The figure uses a constructed example: \emph{Verdier} is an invented name introduced here, not a quotation from a dataset.
It follows one claim wording extracted from a hypothetical model response and shows that its label depends on the prompt and truth contract.
The wording is fixed, but each prompt establishes a different discourse frame, so the two occurrences do not express the same contextualized claim.

\begin{figure}[t]
\centering
\resizebox{0.94\linewidth}{!}{%
\begin{tikzpicture}[x=1cm,y=1cm]
  \node[input,text width=6.7cm] (s) at (0,0)
    {claim extracted from a model response:\\``the caf\'e \emph{Verdier} and its owner existed in 1920s Paris''};
  \node[decision,text width=4.25cm] (qa) at (-3.0,-2.35)
    {\textbf{Prompt:} ``Did caf\'e Verdier and its owner exist in 1920s Paris?''\\[-1pt]{\scriptsize $\TC(p)$\\$O_p=$ historical evidence\\$\kappa_p=0$: factual only}};
  \node[decision,text width=4.25cm] (fi) at (3.0,-2.35)
    {\textbf{Prompt:} ``Write a fictional story set in 1920s Paris about caf\'e Verdier and its owner.''\\[-1pt]{\scriptsize $\TC(p)$\\$O_p=$ prompt constraints\\$\kappa_p=2$: in-frame invention}};
  \node[verdictH] (h) at (-3.0,-4.45) {\Hall};
  \node[verdictLD] (ld) at (3.0,-4.45) {\LD};
  \node[font=\scriptsize,align=center,text width=3.7cm] at (-3.0,-5.10)
    {unsupported real-world assertion};
  \node[font=\scriptsize,align=center,text width=3.7cm] at (3.0,-5.10)
    {invention inside the fictional frame};
  \draw[arrow] (s.south west) ++(0.9,0) -- (qa.north);
  \draw[arrow] (s.south east) ++(-0.9,0) -- (fi.north);
  \draw[arrow] (qa) -- (h);
  \draw[arrow] (fi) -- (ld);
\end{tikzpicture}}
\caption{\textbf{One extracted claim wording, two contract-dependent labels.}
The same constructed claim wording is evaluated under two prompts.
Under the historical-QA prompt it is an unsupported real-world assertion and receives \Hall; under the fiction-writing prompt it is an in-frame invention permitted by $K_p$ and receives \LD.
Strictly, each prompt changes the contextualized claim, so the figure shows contract dependence, not a contradiction about one fixed proposition.
Both labels are claim-level, not verdicts for the complete response.}
\label{fig:flip}
\end{figure}

\paragraph{Evidence and scope}
We support the position in two ways.
First, we map forty hallucination and creativity evaluation resources against the components of the proposed contract.
This purposive, author-coded mapping identifies a pattern in the sample rather than estimating prevalence in the full literature.
Second, we apply the decision rule to five worked cases involving reference evidence, the content-permission policy, epistemic presentation, scope, and $\sigma_p$-conditioned claim typing.
These cases illustrate the intended behavior of the rule but do not establish annotation reliability or completeness.

\paragraph{Contributions}
This position paper makes three contributions.
First, it introduces the truth contract $\TC(p)=(O_p,\sigma_p,K_p)$.
The contract separates reference evidence, requested style level, and the content-permission policy, and distinguishes \Hall\ from \LD.
Second, it maps a sample of factuality and creativity protocols to identify where task permissions and form-conditioned factuality checks remain implicit.
Third, it proposes a research agenda for contract-aware benchmarks and mitigation methods that reduce \Hall\ without suppressing \LD\ or factual stylistic creativity.

\section{Related Evaluation Paradigms}\label{sec:related}

\subsection{Hallucination taxonomies and the missing contract layer}

Hallucination surveys organize the phenomenon along several related but non-equivalent axes.
\citet{ji_survey_2023} distinguish intrinsic hallucinations, which contradict the source, from extrinsic hallucinations, which cannot be verified from it.
\citet{huang_survey_2025} separate factuality hallucinations---factual contradiction or fabrication---from faithfulness hallucinations involving instruction, context, or logical inconsistency.
\citet{zhang_sirens_2025} instead distinguish input-, context-, and fact-conflicting outputs.
Other reviews organize causes across the model lifecycle or examine factuality and faithfulness metrics for open-ended generation \citep{wang_factuality_survey_2024,malin_faithfulness_review_2025,lamba_lifecycle_2026}.
From a creativity perspective, \citet{jiang_survey_2024} distinguish divergent generation from convergent evaluation.

These surveys describe what kind of mismatch occurred, where it may arise, and how it may be measured or used.
They do not, however, jointly represent the relevant evidence, requested style, and permitted divergence as explicit task variables.
Our truth contract adds this complementary decision layer: $O_p$ selects the evidence, $\sigma_p$ describes the requested style, and $K_p=(\kappa_p,\Gamma_p,\rho_p)$ specifies whether evidence-unknown content is permitted, where it is permitted, and how it must be presented.
Descriptive taxonomy and contract-relative label should therefore remain distinct.
As Figure~\ref{fig:survey-contract} shows, an evidence-unknown claim may be described as extrinsic or fact-conflicting, yet receive \Hall\ under a factual-only contract and \LD\ when it falls within the licensed scope and satisfies the required epistemic presentation.

\begin{figure*}[t]
\centering
\resizebox{0.98\linewidth}{!}{%
\begin{tikzpicture}[font=\small]
  \tikzset{
    surveybox/.style={rectangle,rounded corners=4pt,draw=clrB!70,fill=clrB!6,
      text width=6.4cm,align=left,inner sep=8pt,minimum height=4.05cm},
    contractbox/.style={rectangle,rounded corners=4pt,draw=clrP!75,fill=clrP!6,
      text width=6.4cm,align=left,inner sep=8pt,minimum height=4.05cm},
    resultbox/.style={rectangle,rounded corners=3pt,draw=gray!60,fill=gray!5,
      text width=6.4cm,align=center,inner sep=7pt,font=\small\bfseries},
    sharedbox/.style={rectangle,rounded corners=4pt,draw=clrG!70,fill=gray!5,
      text width=7.4cm,align=center,inner sep=7pt,font=\small\bfseries}
  }

  \node[sharedbox] (claim) at (0,0)
    {One extracted claim $c$ (from prompt $p$ and response $y$)};

  \node[surveybox,anchor=north east] (surveys) at (-0.35,-1.35) {%
    \textbf{Existing survey lenses: \emph{describe} the divergence}\\[3pt]
    {\scriptsize
    \textbf{Relation to the source:} intrinsic / extrinsic \citep{ji_survey_2023}\\
    \textbf{Type of inconsistency:} factuality / faithfulness \citep{huang_survey_2025}\\
    \textbf{Conflict target:} input / context / fact \citep{zhang_sirens_2025}\\
    \textbf{Other views:} lifecycle cause, evaluation metric, or creative phase
    \citep{lamba_lifecycle_2026,wang_factuality_survey_2024,malin_faithfulness_review_2025,jiang_survey_2024}}
  };

  \node[contractbox,anchor=north west] (contract) at (0.35,-1.35) {%
    \textbf{Truth-contract lens: \emph{judge} the claim in this task}\\[3pt]
    {\scriptsize
    $O_p$: which evidence supports or contradicts the claim?\\
    $\sigma_p$: what style did the prompt request?\\
    $K_p=(\kappa_p,\Gamma_p,\rho_p)$: is evidence-unknown content allowed, within which scope, and under which presentation?\\[2pt]
    Requested style guides interpretation but never licenses unsupported content.}
  };

  \node[resultbox,below=0.75cm of surveys] (category)
    {Descriptive category\\[-1pt]
    {\scriptsize\mdseries e.g.\ \emph{extrinsic} or \emph{fact-conflicting}}};

  \node[resultbox,below=0.75cm of contract] (label)
    {Contract-relative label\\[-1pt]
    {\scriptsize\mdseries \SUP\ (supported), \LD\ (licensed, evidence-unknown),\\ or \Hall\ (contradicted or unlicensed)}};

  \node[sharedbox,below=0.9cm of $(category.south)!0.5!(label.south)$] (joint)
    {The two lenses are complementary: the same claim keeps \emph{one} descriptive category \emph{and} receives \emph{one} contract-relative label.};

  \draw[arrow] (claim.south west) ++(0.9,0) -- (surveys.north);
  \draw[arrow] (claim.south east) ++(-0.9,0) -- (contract.north);
  \draw[arrow] (surveys.south) -- (category.north);
  \draw[arrow] (contract.south) -- (label.north);
  \draw[arrow] (category.south) -- (joint.north west);
  \draw[arrow] (label.south) -- (joint.north east);
\end{tikzpicture}}
\caption{\textbf{Existing hallucination taxonomies and the complementary truth-contract layer.}
Existing surveys \emph{describe} a claim: its relation to a source, the type or target of an inconsistency, its possible cause, its evaluation, or its role in a creative process.
The truth-contract lens does not replace these descriptions; it \emph{judges} the same claim by adding the task-specific information needed to assign \SUP, \Hall, or \LD.
The two outputs coexist: an evidence-unknown claim can keep one descriptive category (for example, extrinsic or fact-conflicting) while receiving \Hall\ when unlicensed or \LD\ when $K_p$ licenses it within $\Gamma_p$ and its presentation satisfies $\rho_p$.}
\label{fig:survey-contract}
\end{figure*}

\subsection{Adjacent evaluation paradigms}

Faithfulness asks whether a claim is supported by provided evidence, whereas factuality asks whether it is correct against an external reference \citep{maynez_faithfulness_2020,kryscinski_evaluating_2020}.
Both define evidence relations.
The content-permission policy asks a different question: what may an output do when the reference evidence does not entail a claim?

Intent-aware evaluation checks whether an output satisfies query constraints \citep{hao_faithqa_2025}.
This supports the broader view that hallucination evaluation should consider the task rather than the output alone.
However, general intent combines many requirements and does not isolate reference evidence, the content-permission policy, and epistemic presentation.

Recent work distinguishes potentially valuable or ``intelligent'' hallucinations from defective ones \citep{jiang_survey_2024,sui_confabulation_2024,yang_hicbench_2025}.
Our distinction does not rest on value.
Permission is evaluated first: an ingenious fabricated medical fact remains \Hall, whereas an unhelpful but in-frame fictional invention remains \LD\ with low usefulness.

Foregrounding and style-transfer research shows that marked form can be described and manipulated while attempting to preserve content \citep{leech_short_style_1981,vanpeer_stylistics_2020,pavlick_empirical_2016,rao_dear_2018,briakou_evaluating_2021}.
Hallucination evaluation, by contrast, focuses mainly on support and contradiction.
The missing bridge is a protocol that varies stylistic marking while holding the expressed claims and reference evidence fixed, then tests whether extraction and claim-level labels remain stable.

\section{Truth Contracts and Claim-Level Labels}\label{sec:framework}

\subsection{The truth contract}

\textbf{Truth contract (\TC).}
For prompt $p$, the truth contract is
\[
  \TC(p)=(O_p,\sigma_p,K_p), \qquad K_p=(\kappa_p,\Gamma_p,\rho_p).
\]
The three components answer different questions: what supports a claim, what style level the task requests, and what unsupported content the task permits.
We drop the subscript $p$ when the prompt is clear from context.

\textbf{Reference evidence ($O_p$).}
$O_p$ is the reference evidence against which a claim is evaluated, such as a source document, retrieval set, database, gold labels, or a specified external source.
The phrase ``world knowledge'' is only a shorthand; reproducible evaluation requires a concrete evidence source or an explicit adjudication procedure \citep{ji_survey_2023,bang_hallulens_2025}.

\textbf{Content-permission policy ($K_p$).}
$K_p$ specifies three properties:
\begin{itemize}
  \item $\kappa_p\in\{0,1,2\}$: the permission level, where $0$ is factual only, $1$ allows limited and explicitly marked divergence, and $2$ allows invention inside a declared frame;
  \item $\Gamma_p$: the \emph{permission scope}, meaning the topics, entities, or fictional world in which evidence-unknown content may be introduced;
  \item $\rho_p$: the \emph{required epistemic presentation}, meaning how the response must signal the status of that content, for example with a hedge, an explicit hypothesis marker, or a declared fictional frame.
\end{itemize}
The scale orders only the breadth of permitted departure from the reference evidence.
It is not a scale of quality, creativity, or factuality.
In particular, limited speculation and fiction carry different commitments and presentation requirements.

A factual-only task has an empty permission scope: every factual claim must be supported.
For $\Gamma_p$, consider a medical brainstorming task that permits possible diagnoses but not invented drug doses; the diagnosis space is inside the scope, whereas invented treatment content, such as drug doses, is outside it.
For $\rho_p$, ``this may be early Lyme disease, but testing is required'' satisfies a request to mark uncertainty, while ``this is early Lyme disease'' does not.
In fiction, the declared story frame can satisfy $\rho_p$ for invented characters and events, but $\Gamma_p$ may still exclude fabricated real-world claims about named people \citep{searle_logical_1975}.

\subsection{Requested style levels}

The truth contract distinguishes three requested style levels, $\sigma_p\in\{0,1,2\}$:
\begin{itemize}
  \item $\sigma_p=0$ (neutral/literal): minimal rhetorical transformation and no salient authorial voice;
  \item $\sigma_p=1$ (moderately marked): local metaphor, register shift, or a light stylistic voice while the intended claims remain transparent;
  \item $\sigma_p=2$ (strongly marked): sustained figurative language, a salient voice, or substantial stylistic transformation.
\end{itemize}
The levels rank the degree of formal marking.
They do not rank creativity or factuality, and they do not grant permission to invent.
They operationalize \emph{foregrounding}: the prominence produced when a text departs from a neutral background through unexpected deviation or patterned regularity \citep{leech_short_style_1981,vanpeer_stylistics_2020}.
Metaphor is a central semantic case, but foregrounding may also be lexical, syntactic, phonological, or discourse-level.
Surface unexpectedness is therefore not, by itself, evidence of factual error.
Style is multi-dimensional, so this three-level scheme is a working discretization rather than a validated universal scale \citep{kang_style_2021,jin_deep_2022}.
The boundaries, especially between $\sigma_p=1$ and $\sigma_p=2$, require empirical validation.

The requested style level $\sigma_p$ must be distinguished from the observed style level $\widehat\sigma(s,p)$ in an output span $s$.
For example, a prompt can request $\sigma_p=2$ while the model produces neutral prose with $\widehat\sigma(s,p)=0$.
This mismatch is a style-compliance error, not a hallucination.

The diagnostic role of $\sigma_p$ is upstream of factual verification.
The requested style level guides the interpretation of a span, while the observed style level records the form that the model produced.
Claim extractors, entailment systems, and semantic metrics may react differently to literal and figurative realizations of the same claim \citep{chen_menli_2023,aynetdinov_semscore_2024,lai_multidimensional_2023,pauli_mind_2025}.
As a result, stylistic form can change which claims an evaluator extracts and verifies even though $\sigma_p$ does not authorize new content.

\subsection{Unit of analysis and span typing}

We use a claim-level unit because a response may mix supported, unsupported, and non-claim material \citep{min_factscore_2023,bayat_factbench_2024}.
Let $s$ denote a surface span and let $c(s,p)$ denote a truth-conditional claim obtained by interpreting $s$ in the context established by prompt $p$.
The same wording can yield different contextualized claims in fiction, metaphor, and hypothetical language.

Before assigning claim-level labels, the evaluator performs a contract-guided span analysis
\[
\mathcal X_{\sigma_p}(s,p)
=
\bigl(\mathcal C_{\sigma_p}(s,p),\widehat\sigma(s,p),z_{\sigma_p}(s,p)\bigr).
\]
Here, $\mathcal C_{\sigma_p}(s,p)$ is the set of extracted truth-conditional claims.
The observed style level $\widehat\sigma(s,p)\in\{0,1,2\}$ describes the produced form, and $z_{\sigma_p}(s,p)\in\{0,1\}$ marks an independently identifiable stylistic contribution (\SV).
This analysis produces four span types:
\begin{itemize}
  \item \emph{other}: $\mathcal C_{\sigma_p}=\varnothing$ and $z_{\sigma_p}=0$; the span is outside factual and stylistic evaluation;
  \item \emph{SV-only}: $\mathcal C_{\sigma_p}=\varnothing$ and $z_{\sigma_p}=1$; no factual claim is verified;
  \item \emph{claim-only}: $\mathcal C_{\sigma_p}\neq\varnothing$ and $z_{\sigma_p}=0$; each extracted claim is verified;
  \item \emph{mixed}: $\mathcal C_{\sigma_p}\neq\varnothing$ and $z_{\sigma_p}=1$; the span keeps \SV\ while each claim is verified.
\end{itemize}
Metaphor is not automatically removed from factual evaluation.
We annotate a metaphorical span as mixed when adjudicators recover both truth-conditional content and an independently identifiable stylistic contribution.
Some metaphors also introduce implicatures or new claims; those claims must be extracted and verified.

To compare styles, expert annotators first restate each claim as a neutral paraphrase.
We call two claims \emph{canonical matches} when annotators judge that these paraphrases express the same factual commitment.
We write $\mathcal C^{*}$ for a claim set after this matching, and $\equiv$ for equality between canonical claim sets.
We assess consistency at two stages: recovery of canonical claims and assignment of claim-level labels.
Consider expert-adjudicated realizations $s_0,s_1,s_2$ produced under contracts that differ only in requested style level.
If the realizations preserve the same claims, an evaluator should recover equivalent canonical claim sets and stable claim-level labels:
\[
\begin{aligned}
\mathcal C^{*}_{\sigma_{p_0}}(s_0,p_0)
&\equiv \mathcal C^{*}_{\sigma_{p_1}}(s_1,p_1)\\
&\equiv \mathcal C^{*}_{\sigma_{p_2}}(s_2,p_2),
\end{aligned}
\qquad
\begin{aligned}
V(c\mid p_0)&=V(c\mid p_1)\\
&=V(c\mid p_2),
\end{aligned}
\]
where $\sigma_{p_0}=0$, $\sigma_{p_1}=1$, and $\sigma_{p_2}=2$.
The observed style levels and \SV\ flags may legitimately differ.
An automatic evaluator fails if it drops a recoverable claim as style, literalizes stylistic material into a spurious claim, or changes the claim-level label after the canonical claim has been matched.
This requirement applies only to claim-preserving variants.
For models, the target capability is to produce stylistically creative yet factually grounded output whenever $K_p$ requires grounding.
Failure to produce the requested style level is an instruction-following error, whereas adding an unsupported claim under a grounded-only $K_p$ is a truth-contract violation.

\subsection{Evidence state and claim label}

For every truth-conditional claim $c$, record the evidence relation
\[
E(c,O_p)\in\{\textsc{entailed},\textsc{contradicted},\textsc{unknown}\}.
\]
The last value means that the available evidence neither supports nor contradicts the claim.
Evidence state alone does not determine the contractual claim label.

Let $L(c,K_p)$ indicate whether permission level $\kappa_p$ permits claim $c$ within scope $\Gamma_p$, and let $M(c,s,K_p)$ indicate whether its realization in span $s$ satisfies the epistemic-presentation requirement $\rho_p$.
Then
\[
V(c\mid p)=
\begin{cases}
\SUP & \text{if }E(c,O_p)=\textsc{entailed},\\[1mm]
\LD & \text{if }E(c,O_p)=\textsc{unknown},\ L=1,\ M=1,\\[1mm]
\Hall & \text{otherwise.}
\end{cases}
\]
\SUP\ means \emph{supported}: $O_p$ entails the claim.
\Hall\ means \emph{hallucination}: the claim is contradicted, or it is evidence-unknown without the required permission, scope, or presentation.
\LD\ means \emph{licensed divergence}: $O_p$ does not establish the claim, but $K_p$ explicitly allows that type of content inside $\Gamma_p$ and under $\rho_p$.
An \LD\ label records contractual permission, not proof that the claim is true in the external world.

By default, a claim that contradicts $O_p$ receives \Hall.
For fiction, $O_p$ includes the declared fictional frame and its constraints; disagreement with the real world alone does not make an in-frame invention contradicted.
If an entailed claim violates a requested hedge or style, it remains \SUP\ but is additionally flagged with a task-compliance error.
Usefulness $U(c,p)$ is reported as a separate task-dependent score over \LD\ claims.
A low usefulness score does not change \LD\ into \Hall.

An explicit refusal or a statement such as ``the available evidence is insufficient'' is evaluated as a meta-claim about the evidence.
It is not automatically treated as asserting the embedded unsupported claim.
This distinction prevents calibrated abstention from being mislabeled as hallucination.

\paragraph{Claim labels versus response-level assessment}
The rule above returns one claim-level label $V(c_i\mid p)$ for each extracted claim $c_i$ in a response $y$.
It does not by itself return a verdict for the complete response.
Current evaluation methods often aggregate claim-level results with a proportion, mean score, or binary rule such as whether any error occurs \citep{min_factscore_2023,bayat_factbench_2024}.
These aggregations answer different questions: one severe \Hall\ may matter more than several minor supported claims, while a creative response can contain no \Hall\ yet still provide too little useful \LD.
We write $R(y\mid p)$ for a future contract-aware response-level assessment.
This paper does not define an optimal aggregation function for $R$; such a function should state how it combines claim labels with coverage, severity, usefulness, and style compliance under $\TC(p)$.

Figure~\ref{fig:decision} summarizes the procedure.

\begin{figure*}[t]
\centering
\resizebox{0.95\linewidth}{!}{%
\begin{tikzpicture}[x=1cm,y=1cm]
  \node[input,text width=5.7cm] (response) at (0,0) {model response ($y$)};
  \node[decision,text width=6.2cm] (contract) at (0,-1.35)
    {read the truth contract $\TC(p)=(O_p,\sigma_p,K_p)$\\[-1pt]
     {\scriptsize with $K_p=(\kappa_p,\Gamma_p,\rho_p)$}};
  \node[decision,text width=6.2cm] (label) at (0,-2.70)
    {label each span $s$ as a factual claim, stylistic variation (\SV), both, or neither};
  \node[decision,text width=5.6cm] (hasclaim) at (0,-4.05)
    {does span $s$ express a factual claim $c$?};
  \node[verdict,text width=4.4cm,align=center] (styleonly) at (6.4,-4.05)
    {retain the \SV\ label;\\no factuality check};
  \node[decision,text width=5.6cm] (plainclaim) at (0,-5.40)
    {recover a canonical claim $c^*$ in clear, literal wording};
  \node[decision,text width=5.6cm] (evidence) at (0,-6.75)
    {compare $c^*$ with the reference evidence $O_p$};
  \node[verdict,text width=3.5cm,align=center] (sup) at (-6.0,-6.75)
    {supported claim\\(\SUP)};
  \node[verdictH,text width=3.5cm,align=center] (contradicted) at (6.0,-6.75)
    {hallucination\\(\Hall)};
  \node[decision,text width=6.2cm] (permission) at (0,-8.25)
    {if evidence is unknown: does $K_p$ allow $c^*$ at level $\kappa_p$, inside $\Gamma_p$, and under $\rho_p$?};
  \node[verdictLD,text width=3.8cm,align=center] (licensed) at (-3.5,-9.75)
    {licensed divergence\\(\LD)};
  \node[verdictH,text width=3.8cm,align=center] (unlicensed) at (3.5,-9.75)
    {hallucination\\(\Hall)};
  \node[decision,dashed,draw=gray!70,text width=9.2cm] (responselevel) at (0,-11.35)
    {after all claims: combine their labels into a response-level assessment $R(y\mid p)$\\[-1pt]
     {\scriptsize no optimal contract-aware aggregation rule is defined yet}};
  \draw[arrow] (response) -- (contract);
  \draw[arrow] (contract) -- (label);
  \draw[arrow] (label) -- (hasclaim);
  \draw[arrow] (hasclaim) -- node[edge,left]{yes} (plainclaim);
  \draw[arrow] (hasclaim.east) -- node[edge,above]{no} (styleonly.west);
  \draw[arrow] (plainclaim) -- (evidence);
  \draw[arrow] (evidence.west) -- node[edge,above]{entailed} (sup.east);
  \draw[arrow] (evidence.east) -- node[edge,above]{contradicted} (contradicted.west);
  \draw[arrow] (evidence) -- node[edge,left]{unknown} (permission);
  \draw[arrow] (permission.south west) -- node[edge,above,pos=0.62]{yes} (licensed.north);
  \draw[arrow] (permission.south east) -- node[edge,above,pos=0.62]{no} (unlicensed.north);
\end{tikzpicture}}
\caption{\textbf{Plain-language span-to-claim evaluation procedure.}
The evaluator reads the truth contract $\TC(p)$ and labels the role of each span $s$.
Stylistic variation (\SV) without a factual claim bypasses factual verification.
Each recovered canonical claim $c^*$ is compared with $O_p$; when the evidence is unknown, $K_p=(\kappa_p,\Gamma_p,\rho_p)$ determines whether the claim receives \LD\ or \Hall.
The resulting \SUP, \Hall, and \LD\ labels remain claim-level outputs.
Combining them into a response-level assessment $R(y\mid p)$ is shown as a separate open problem.}
\label{fig:decision}
\end{figure*}

\subsection{Representative contracts}

Table~\ref{tab:contracts} applies the rule to representative prompts.
Each card specifies one contract and evaluates one or two claims recovered from example output spans.

\begin{table*}[t]
\centering
\small
\caption{\textbf{The thesis in one view.}
Each colored card fixes one truth contract $\TC(p)=(O_p,\sigma_p,K_p)$, with $K_p=(\kappa_p,\Gamma_p,\rho_p)$, and lists one or two evaluated claims from example outputs.
\SUP, \Hall, and \LD\ are claim-level labels and \SV\ is an auxiliary span label; the table assigns no verdict to a complete response.}
\label{tab:contracts}
\setlength{\fboxsep}{5pt}
\fcolorbox{clrB}{clrB!3}{%
\begin{minipage}{0.94\linewidth}
\textbf{Prompt:} ``Summarize this medical report''\\[2pt]
\null\hfill\colorbox{clrB!12}{\strut $\sigma_p=0,\ \kappa_p=0$}\\[2pt]
\textbf{Truth contract:} $O_p=$ provided report; $\Gamma_p=\varnothing$; $\rho_p=$ factual assertion.\par\smallskip
\begin{tabularx}{\linewidth}{@{}YP{0.12\linewidth}P{0.30\linewidth}@{}}
\textbf{Response claim:} ``The patient has diabetes'' (not in the report) & \textcolor{clrH}{\bfseries \Hall} & unsupported by $O_p$ and not licensed\\
\end{tabularx}
\end{minipage}}

\smallskip
\fcolorbox{clrLD}{clrLD!3}{%
\begin{minipage}{0.94\linewidth}
\textbf{Prompt:} ``Explain photosynthesis like a poet, but keep the science accurate''\\[2pt]
\null\hfill\colorbox{clrLD!12}{\strut $\sigma_p=2,\ \kappa_p=0$}\\[2pt]
\textbf{Truth contract:} $O_p=$ scientific references; $\Gamma_p=\varnothing$; $\rho_p=$ factual assertion.\par\smallskip
\begin{tabularx}{\linewidth}{@{}YP{0.12\linewidth}P{0.30\linewidth}@{}}
\textbf{Response claim:} ``Chlorophyll catches sunlight, fueling the plant's sugar-making work'' & {\bfseries \SUP+\SV} & supported claim in marked form\\
\addlinespace[3pt]
\textbf{Response claim:} ``Plants make sugar from moonlight'' & \textcolor{clrH}{\bfseries \Hall} & contradicted claim; $\sigma_p=2$ does not license it\\
\end{tabularx}
\end{minipage}}

\smallskip
\fcolorbox{clrP}{clrP!3}{%
\begin{minipage}{0.94\linewidth}
\textbf{Prompt:} ``List possible causes; mark anything unestablished''\\[2pt]
\null\hfill\colorbox{clrP!12}{\strut $\sigma_p=0,\ \kappa_p=1$}\\[2pt]
\textbf{Truth contract:} $O_p=$ clinical evidence; $\Gamma_p=$ possible diagnoses; $\rho_p=$ explicit uncertainty marker.\par\smallskip
\begin{tabularx}{\linewidth}{@{}YP{0.12\linewidth}P{0.30\linewidth}@{}}
\textbf{Response claim:} ``This could be early Lyme disease, but testing is required'' & \textcolor{clrLD}{\bfseries \LD} & unknown, in scope, and properly hedged\\
\addlinespace[3pt]
\textbf{Response claim:} ``This is early Lyme disease'' & \textcolor{clrH}{\bfseries \Hall} & required hedge is missing\\
\end{tabularx}
\end{minipage}}

\smallskip
\fcolorbox{clrO}{clrO!3}{%
\begin{minipage}{0.94\linewidth}
\textbf{Prompt:} ``Write a story set in 1920s Paris''\\[2pt]
\null\hfill\colorbox{clrO!12}{\strut $\sigma_p=1,\ \kappa_p=2$}\\[2pt]
\textbf{Truth contract:} $O_p=$ prompt and historical constraints; $\Gamma_p=$ fictional world; $\rho_p=$ declared fictional frame.\par\smallskip
\begin{tabularx}{\linewidth}{@{}YP{0.12\linewidth}P{0.30\linewidth}@{}}
\textbf{Response claim:} ``The fictional caf\'e Verdier opened at dawn'' & \textcolor{clrLD}{\bfseries \LD} & invention inside $\Gamma_p$\\
\addlinespace[3pt]
\textbf{Response claim:} \emph{A fabricated statement about a real named person, presented as biography} & \textcolor{clrH}{\bfseries \Hall} & real-world claim outside $\Gamma_p$\\
\end{tabularx}
\end{minipage}}
\end{table*}

\section{Purposive Mapping of a Diagnostic Gap}\label{sec:audit}

\subsection{Purposive mapping protocol}

We examined forty author-selected evaluation resources covering factual question answering, grounded generation, intent evaluation, creative writing, divergent thinking, and constrained problem solving.
The mapping is purposive rather than systematic.
It asks whether prominent examples from factuality and creativity research explicitly represent the variables in the proposed contract.

For each resource, we recorded five questions:
\begin{enumerate}
  \item Does the protocol specify a reference evidence source?
  \item Does it vary or score the requested style level while controlling claim content?
  \item Does it vary or score the permission to introduce unsupported content?
  \item Does it represent the scope of that permission?
  \item Does it score the required epistemic presentation?
\end{enumerate}
For the second question, we also recorded whether marked form is evaluated as a possible source of factuality-measurement error.

Appendix~\ref{app:mapping} expresses these observations in the framework's terms.
Each resource receives a coarse task-level code for its reference evidence $O$, its dominant requested style level $\sigma$, and its inferred permission level $\kappa$.
We use $\kappa=0$ for factual-only tasks, $\kappa=1$ for limited divergence constrained by explicit marking, feasibility, tests, or soundness, and $\kappa=2$ for invention inside a declared task frame.
An ``implicit'' code means that the level is inferred from the task design rather than represented as an independent evaluation variable.
The level estimates the breadth of permitted departure; it does not measure benchmark quality or replace the scope and presentation components of $K$.

The mapping was coded in a single pass by one author, so the inferred task profiles are interpretive annotations rather than measurements reported by the original resources.
Independent recoding and a public codebook would be required before treating it as a validated resource, and our conclusions are limited to this sample.
Appendix~\ref{app:mapping} lists the resources and their coarse inferred profiles.

\subsection{Observed pattern in the sample}

The complete resource-level coding appears in Appendix Tables~\ref{tab:full-mapping-factual-a}--\ref{tab:full-mapping-creative}; the following patterns state the main observation directly.

\paragraph{Factuality cluster: low $\sigma$, low $\kappa$}
The twenty-four strict factuality and faithfulness resources generally fix reference evidence, assume neutral wording ($\sigma=0$), and treat unsupported claims as errors under an inferred $\kappa=0$ policy.
The two closest intent or hallucination-taxonomy resources represent part of the problem, but do not jointly score evidence, permission scope, and epistemic presentation.

\paragraph{Creativity cluster: higher $\sigma$ or higher $\kappa$, but rarely both as explicit controls}
Creative-writing resources more often request or reward marked language ($\sigma=1$ or $2$) and imply $\kappa=2$.
They generally treat the permission policy $K=(\kappa,\Gamma,\rho)$ as part of the task rather than as an independently varied evaluation variable.
Divergent-thinking resources also imply $\kappa=2$ but often do not control linguistic style.
The five constrained creative-problem-solving resources instead combine mostly neutral form with a partial $\kappa=1$ policy based on feasibility, tests, or correctness.

\paragraph{The apparent contradiction is a coverage gap}
The two clusters do not provide conflicting measurements of one capability; they evaluate different regions of the truth-contract space.
The main missing region is a controlled crossing of $\sigma\in\{0,1,2\}$ with a fixed grounded-only policy, especially strongly marked but factually grounded language.
Within the mapped sample, no resource represents $O_p$, $\sigma_p$, and $K_p$ as independently varied evaluation variables.
Likewise, none jointly evaluates $\sigma_p$-conditioned span typing, \SV\ placement, extracted claim sets, and downstream claim-level labels.

No factuality or faithfulness resource in the sample treats marked form as a possible source of factuality-measurement error, which answers the second recorded sub-question in the negative.
The sample also underrepresents strongly marked language under a strict factual contract.
Existing resources more often pair factual contracts with neutral language and creative contracts with marked language.
The underrepresented combination corresponds to prompts such as ``explain this scientific result like a poet, but keep every claim accurate.''
Such tasks can reveal whether a factuality verifier mistakes figurative wording for unsupported claim content.
Modern verification pipelines make this a plausible failure mode because claim decomposition, natural-language inference, and semantic similarity may all depend on the surface realization \citep{chen_menli_2023,aynetdinov_semscore_2024}.
This is a measurement hypothesis, not a prevalence result established by the present paper.

Figure~\ref{fig:profiles} organizes the target task profiles by requested style and breadth of permitted departure.

\begin{figure*}[t]
\centering
\resizebox{0.96\linewidth}{!}{%
\begin{tikzpicture}[x=1cm,y=1cm,font=\small]
  \def\cw{3.55}
  \def\rh{1.95}
  \fill[clrB!6] (0,0) rectangle (\cw,\rh);
  \fill[clrB!4] (\cw,0) rectangle (2*\cw,\rh);
  \fill[clrB!3] (2*\cw,0) rectangle (3*\cw,\rh);
  \fill[gray!5] (0,\rh) rectangle (\cw,2*\rh);
  \fill[gray!5] (\cw,\rh) rectangle (2*\cw,2*\rh);
  \fill[gray!7] (2*\cw,\rh) rectangle (3*\cw,2*\rh);
  \fill[clrLD!7] (0,2*\rh) rectangle (\cw,3*\rh);
  \fill[clrLD!8] (\cw,2*\rh) rectangle (2*\cw,3*\rh);
  \fill[clrLD!10] (2*\cw,2*\rh) rectangle (3*\cw,3*\rh);
  \draw[gray!60] (0,0) rectangle (3*\cw,3*\rh);
  \draw[gray!35] (\cw,0)--(\cw,3*\rh);
  \draw[gray!35] (2*\cw,0)--(2*\cw,3*\rh);
  \draw[gray!35] (0,\rh)--(3*\cw,\rh);
  \draw[gray!35] (0,2*\rh)--(3*\cw,2*\rh);
  \draw[very thick,clrH] (2*\cw,0) rectangle (3*\cw,\rh);
  \node[align=center,text width=3.15cm] at (0.5*\cw,0.5*\rh)
    {\textbf{Grounded reporting}\\{\scriptsize paraphrase; no new claim}};
  \node[align=center,text width=3.15cm] at (1.5*\cw,0.5*\rh)
    {\textbf{Styled\\explanation}\\{\scriptsize local marking; claims grounded}};
  \node[align=center,text width=3.15cm] at (2.5*\cw,0.5*\rh)
    {\textbf{Strongly marked\\factual content}\\{\scriptsize sustained style; claims grounded}};
  \node[align=center,text width=3.15cm] at (0.5*\cw,1.5*\rh)
    {\textbf{Flagged\\hypotheses}\\{\scriptsize literal provisional statements}};
  \node[align=center,text width=3.15cm] at (1.5*\cw,1.5*\rh)
    {\textbf{Styled\\hypotheses}\\{\scriptsize provisional content; local marking}};
  \node[align=center,text width=3.15cm] at (2.5*\cw,1.5*\rh)
    {\textbf{Figurative hypotheses}\\{\scriptsize provisional content; sustained style}};
  \node[align=center,text width=3.15cm] at (0.5*\cw,2.5*\rh)
    {\textbf{Plain-form ideation}\\{\scriptsize invention inside task constraints}};
  \node[align=center,text width=3.15cm] at (1.5*\cw,2.5*\rh)
    {\textbf{Styled ideation}\\{\scriptsize invention with moderate marking}};
  \node[align=center,text width=3.15cm] at (2.5*\cw,2.5*\rh)
    {\textbf{Creative writing}\\{\scriptsize strongly marked in-frame invention}};
  \node[font=\scriptsize,align=center] at (0.5*\cw,-0.42){$\sigma_p=0$\\neutral / literal};
  \node[font=\scriptsize,align=center] at (1.5*\cw,-0.42){$\sigma_p=1$\\moderately marked};
  \node[font=\scriptsize,align=center] at (2.5*\cw,-0.42){$\sigma_p=2$\\strongly marked};
  \node[font=\small] at (1.5*\cw,-1.15){requested style level $\sigma_p$};
  \node[font=\scriptsize,anchor=east] at (-0.16,0.5*\rh){$\kappa_p=0$: factual only};
  \node[font=\scriptsize,anchor=east] at (-0.16,1.5*\rh){$\kappa_p=1$: limited divergence};
  \node[font=\scriptsize,anchor=east] at (-0.16,2.5*\rh){$\kappa_p=2$: in-frame invention};
\end{tikzpicture}}
\caption{\textbf{Conceptual task-design space.}
The grid crosses three content-permission levels with the three requested style levels, $\sigma_p\in\{0,1,2\}$.
Each cell is an illustrative task profile, not an empirical frequency or quality ranking.
A higher $\kappa_p$ broadens the type of departure that a task may permit, but scope and presentation constraints still apply.
A higher $\sigma_p$ changes the requested style level but does not license unsupported claims.
Background shading only groups rows by $\kappa_p$ and carries no ordering; the highlighted $\sigma_p=2,\ \kappa_p=0$ cell (strongly marked but grounded) is the region most underrepresented in our sample.}
\label{fig:profiles}
\end{figure*}

\subsection{Why unsupported generation is not the target}

Formal work suggests that some errors remain possible for sufficiently general models operating with incomplete evidence \citep{kalai_calibrated_2024,xu_hallucination_2024,banerjee_llms_2024}.
This does not imply a constant deployment error rate \citep{suzuki_negligible_2025}.
For our position, the main point is operational: suppressing every departure from observed evidence would also suppress hypotheses and in-frame invention.
Evaluation should instead target three outcomes: reducing unlicensed departures (\Hall), preserving licensed ones (\LD), and retaining supported content at the requested $\sigma_p$ level \citep{banerjee_does_2025,sun_hallucinating_2024,he_shakespearean_2025}.

\section{Worked Cases}\label{sec:cases}

The following cases illustrate how one or more contract components affect the analysis.
They are conceptual demonstrations rather than measurements.
Case A uses a documented summarization pattern; Cases B--E are constructed contrasts involving frame, permission, scope, presentation, and style-conditioned claim typing.

\paragraph{Case A: changing the reference evidence}
In XSum, an extrinsic but factual summary statement may be absent from the source document while being correct against external evidence \citep{maynez_faithfulness_2020}.
With $O=$ source document and grounded-only $K$, the statement is unsupported and receives \Hall.
With $O=$ an external evidence set that entails the same claim, it receives \SUP.
The surface text does not change; the reference evidence does.

\paragraph{Case B: changing the discourse frame}
The caf\'e sentence in Figure~\ref{fig:flip} is \Hall\ when interpreted as a real-world historical assertion under a grounded historical-QA contract.
Under a declared fictional frame, the corresponding fictional-world claim is in scope and receives \LD.
The example holds the wording fixed while changing its contextualized claim and content-permission policy.

\paragraph{Case C: changing the epistemic presentation}
Suppose the available evidence cannot determine whether a tool supports several input formats.
Under a grounded contract, presenting support as an established fact receives \Hall.
Under a hypothesis-generation contract, a qualified statement such as ``the tool may support several formats, but this requires version-specific verification'' can receive \LD.
The canonical claim is preserved, but the content-permission policy and epistemic presentation differ.

\paragraph{Case D: testing a scope boundary}
A fictional frame does not license every false statement placed inside a story.
If a response attributes a fabricated biographical claim to a real named person and presents it as real history, the claim falls outside the fictional scope and remains \Hall.
Permission is limited by $\Gamma_p$.

\paragraph{Case E: changing claim type under requested style level}
Under $\sigma_p=0$, ``chlorophyll absorbs sunlight that powers sugar production'' has $\widehat\sigma=0$ and is analyzed as claim-only: its supported claims receive \SUP\ and $z_{\sigma_p}=0$.
Under $\sigma_p=2$, ``chlorophyll catches sunlight, fueling the plant's sugar-making work'' has $\widehat\sigma=2$ and is analyzed as mixed: the same canonical claims remain supported, while the marked realization receives $z_{\sigma_p}=1$ and is reported as \SUP+\SV.
If an evaluator extracts a literal pseudo-claim from the figurative wording, a subsequent \Hall\ label reflects an extraction error, not a factual defect in the intended claim.

Together, Cases A--E isolate the role of reference evidence, discourse frame, epistemic presentation, permission scope, and style-conditioned claim recovery.

\section{Research Agenda}\label{sec:agenda}

The framework motivates six connected research priorities.

\textbf{1. Contract annotation, inference, and reliability.}
Benchmarks should annotate the reference evidence $O_p$, the three-level requested style level $\sigma_p$, the observed style level $\widehat\sigma$, permission level $\kappa_p$, scope, and epistemic presentation separately.
A public codebook should define contextualized claims, claim-preserving style variation, and evidence states.
Reliability should be reported for each layer, and annotator disagreement should be preserved when prompts are ambiguous.
Contract inference must also resolve conflicts among user prompts, system instructions, domain rules, and safety policies.
Multi-turn studies should track when a clarification, correction, or role instruction changes $\TC(p)$ instead of treating the first prompt as a fixed contract for the whole dialogue.

\textbf{2. Factorial benchmark design.}
A diagnostic benchmark should cross $\kappa_p\in\{0,1,2\}$ with $\sigma_p\in\{0,1,2\}$ while holding canonical claims as stable as possible.
Expert-adjudicated triplets should express the same claim in neutral, moderately marked, and strongly marked forms.
Within each cell, targeted contrasts should vary $\Gamma_p$ and $\rho_p$ to distinguish permission breadth from permission scope and required presentation.
Evaluation should report span type, observed-style accuracy, \SV\ placement, claim-set equivalence, evidence state, and claim-level label.
A successful evaluator need not produce identical raw extractions across styles; it should recover equivalent canonical claims and stable labels after matching.
The mapping in Appendix~\ref{app:mapping} provides possible comparators, not a validated benchmark.

\textbf{3. Style-robust claim recovery and verification.}
Claim extractors and verifiers should be stress-tested on metaphor, analogy, irony, narrative voice, hypotheses, and declared fiction.
Experiments should separate span-type errors, claim-recovery errors, evidence errors, and contract-label errors so that a final \Hall\ label can be traced to the stage that produced it.
Comparisons across extraction, natural-language-inference, retrieval, and LLM-judge pipelines should report both accuracy and calibration under each $\sigma_p$ level.
The central test is whether matched canonical claims keep stable evidence states and labels when only their permitted surface realization changes.

\textbf{4. Response-level aggregation.}
A complete response $y$ requires a response-level assessment $R(y\mid p)$ built from its claim-level labels and other task outcomes.
Research should compare transparent aggregation rules, including the proportion of \Hall\ claims, a worst-case rule, severity-weighted scoring, and task-utility-aware scoring.
Each aggregate should be reported together with the underlying claim-label distribution, claim coverage, style compliance, and sensitivity to the chosen rule.
Different contracts may justify different aggregations; for example, one severe unsupported medical claim and one harmless fictional detail should not be treated as interchangeable events.
We leave the design and validation of an optimal contract-aware $R$ open rather than building it into the claim-level definition.

\textbf{5. Mitigation that preserves licensed divergence and factual style.}
Retrieval, verification, abstention, and constrained decoding should be assessed on three outcomes: reducing \Hall\ under grounded contracts, preserving useful \LD\ under permissive contracts, and preserving supported content at $\sigma_p=1$ or $2$.
Usefulness should remain a separate score with its own uncertainty \citep{acar_creativity_2019,marco_reader_2025}.
Results should expose trade-offs among these outcomes, for example through per-contract scores or Pareto comparisons, rather than through one aggregate hallucination rate.
This would test whether a mitigation method makes a model more reliable without making it uniformly literal, overly cautious, or less useful in creative tasks.

\textbf{6. Generalization across domains, languages, and interactions.}
The three-level $\sigma_p$ and $\kappa_p$ scales should be tested in high-stakes factual domains, educational explanation, journalism, scientific communication, brainstorming, and fiction.
Multilingual work should examine whether style markers, hedges, and genre conventions lead to comparable annotations across languages and cultures.
Interactive studies should also measure whether users understand the model's epistemic presentation and whether contract-aware outputs improve trust calibration and task utility.

\section{Discussion}\label{sec:discussion}

\subsection{Evaluation implications}

Contract-aware evaluation separates claim-level labels from response-level assessment.
At claim level, a departure is supported, hallucinated, or licensed.
At span level, \SV\ is an orthogonal flag: it may occur alone or accompany any claim-level label when stylistic material and claim content coexist.
At response level, $R(y\mid p)$ must summarize a possibly mixed set of \SUP, \Hall, and \LD\ labels together with task-level properties.
A response may still be too conservative for a creative task even when it contains no \Hall.
Likewise, a response may remain supported but reduce a requested $\sigma_p=2$ style to neutral prose.
These are output-level task-compliance failures, not new claim labels.
Conversely, satisfying $\sigma_p=2$ does not excuse a fabricated claim under a grounded-only $K_p$.
Reporting claim-level labels, their aggregation rule, and style compliance separately makes ``creative and factual when required'' an evaluable target.

The framework also separates the claim label from severity.
A false drug dose and a wrong trivia date can both receive \Hall, but their costs differ.
Severity should weight the penalty after the contractual claim label rather than determine whether a violation occurred.

\subsection{Scope and validation priorities}

This position paper presents an operational starting point and makes its validation path explicit.
First, contract specification necessarily includes normative choices about the relevant evidence, permission boundary, and authority of user, system, safety, legal, and domain instructions.
Making these choices explicit is a feature of the framework; future multi-turn work can extend the prompt-centered formulation to contracts that change during dialogue.

Second, the three-level $\sigma_p$ and $\kappa_p$ scales are working operational scales intended to support testable comparisons.
Human annotation studies can refine their thresholds while $\Gamma_p$ and $\rho_p$ preserve task-specific distinctions that a single ordinal level cannot express.
Metaphor, irony, hypothesis, and fiction provide demanding but informative tests for claim extraction, \LD\ annotation, and style robustness.
The claim-level rule is therefore specified now, while alternative response-level aggregation rules $R(y\mid p)$ remain open for direct empirical comparison.

Third, the purposive, author-coded mapping establishes a concrete coverage pattern in the selected resources.
A larger systematic review, independent recoding, and a public codebook would strengthen estimates of how often each truth-contract profile occurs in the broader literature.
These boundaries define the next validation steps, and they leave the paper's central distinction fully testable: evidential support, content permission, and requested style.

\section{Conclusion}\label{sec:conclusion}

Hallucination is not a context-free property of a sentence.
It is a claim-level label relative to a task's reference evidence, requested style level, and content-permission policy.
The truth contract $\TC(p)=(O_p,\sigma_p,K_p)$ separates these components.
An evidence-unknown claim is \LD\ only when it falls within the permitted scope and uses the required epistemic presentation; otherwise it is \Hall.
Usefulness, severity, and style compliance remain separate.
A complete response may contain several labels, and its response-level assessment depends on an aggregation rule that this position paper deliberately leaves open.

The purposive mapping suggests that these variables are rarely represented together in the resources we examined.
The worked cases show why each variable matters.
The next step is to validate contract-guided span typing, measure label stability after claim matching, design transparent response-level aggregation, and test whether mitigation can reduce \Hall\ while preserving \LD\ and factually grounded style.

\appendix

\appendixsection{Additional Diagnostics and Response-Level Aggregation}%
\makeatletter\def\@currentlabel{A}\makeatother\label{app:diagnostics}

Section~\ref{sec:framework} defines the truth contract, span types, and claim-label rule.
To avoid repeating that formalism, this appendix records only additional diagnostics needed for an empirical implementation.

\begin{itemize}
  \item \emph{Span-type error}: incorrect assignment of other, SV-only, claim-only, or mixed.
  \item \emph{Claim-recovery error}: a canonical claim is dropped, added spuriously, or changed during neutral paraphrase.
  \item \emph{Claim-label instability}: claim-matched variants receive different labels when $O_p$ and $K_p$ are fixed.
  \item \emph{Aggregation sensitivity}: the response-level ranking changes when the evaluator changes the rule used to combine claim labels.
\end{itemize}

The first three diagnostics require a benchmark distribution and expert-adjudicated claim matches before they can be estimated as error rates.
Claim-label instability is an error only when both $O_p$ and $K_p$ remain fixed.
When $K_p$ changes, the same evidence-unknown claim may intentionally change from \Hall\ to \LD; this is contract dependence, not relative truth.

For a complete response $y$, an evaluation should publish the number or proportion of \SUP, \Hall, and \LD\ claims, the observed style compliance, and any severity or usefulness weights before reporting an aggregate $R(y\mid p)$.
Because the present paper does not validate one aggregation rule, comparisons should report sensitivity to plausible alternatives rather than present $R$ as a uniquely determined final verdict.

\appendixsection{Resource Mapping}%
\makeatletter\def\@currentlabel{B}\makeatother\label{app:mapping}

Tables~\ref{tab:full-mapping-factual-a}--\ref{tab:full-mapping-creative} list the forty resources used in the purposive mapping.
Each row encodes the reference evidence $O$, the dominant requested style level $\sigma$, and the inferred permission level $\kappa$.
The $\sigma$ column uses the paper's three levels; ``not controlled'' means that style is neither fixed nor scored as a distinct variable.
The $\kappa$ column uses $0$ for factual only, $1$ for limited divergence, and $2$ for in-frame invention.
``Implicit'' means that the level is inferred from the task design, while ``partial'' means that the resource encodes a constraint such as correctness or feasibility without representing the full policy $(\kappa,\Gamma,\rho)$.
Thus, an implicit $\kappa=2$ code does not mean that a benchmark explicitly annotates the content-permission policy; it means that invention is needed or accepted by the task although its permission boundary is not independently evaluated.
The task profiles are coarse author annotations inferred from the original protocols, not labels reported by those resources.
They require a public codebook and independent recoding before release as a validated dataset.

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{0.98}
\caption{\textbf{Factuality and faithfulness resources in the purposive mapping (part I).}
The recurring profile is fixed reference evidence, implicit $\sigma=0$, and inferred factual-only permission ($\kappa=0$).
No listed resource crosses a grounded-only policy with controlled $\sigma=1$ or $2$.
These are author-coded profiles, not labels reported by the original resources.}
\label{tab:full-mapping-factual-a}
\begin{tabularx}{0.98\linewidth}{@{}P{0.23\textwidth}P{0.18\textwidth}P{0.20\textwidth}P{0.11\textwidth}Y@{}}
\toprule
\textbf{Resource} & \textbf{Task} & \textbf{$O$: reference evidence or rubric} & \textbf{Inferred $\sigma$} & \textbf{Inferred $\kappa$ / status}\\
\midrule
\multicolumn{5}{l}{\textbf{Strict factuality and faithfulness}}\\
FEVER \citep{thorne_fever_2018} & Fact checking & Wikipedia evidence & $0$ (implicit) & $0$ (implicit)\\
TruthfulQA \citep{lin_truthfulqa_2022} & Factual QA & External factual reference & $0$ (implicit) & $0$ (implicit)\\
PopQA \citep{mallen_when_2023} & Long-tail QA & Wikidata & $0$ (implicit) & $0$ (implicit)\\
SimpleQA \citep{wei_measuring_2024} & Factual QA & External factual reference & $0$ (implicit) & $0$ (implicit)\\
TriviaQA \citep{joshi_triviaqa_2017} & Reading QA & Evidence documents & $0$ (implicit) & $0$ (implicit)\\
DefAn \citep{rahman_defan_2024} & Factual QA & External factual reference & $0$ (implicit) & $0$ (implicit)\\
HaluEval \citep{li_halueval_2023} & QA and summarization & World/reference context & $0$ (implicit) & $0$ (implicit)\\
FactBench \citep{bayat_factbench_2024} & Factual QA & Retrieved evidence & $0$ (implicit) & $0$ (implicit; undecidable retained)\\
HalluLens \citep{bang_hallulens_2025} & Factual QA & Task-specific evidence & $0$ (implicit) & $0$ (implicit)\\
FRANK \citep{pagnoni_understanding_2021} & Summarization & Source documents & $0$ (implicit) & $0$ (implicit)\\
QAGS \citep{wang_asking_2020} & Summarization & Source documents & $0$ (implicit) & $0$ (implicit)\\
SummEval \citep{fabbri_summeval_2021} & Summarization & Source documents & $0$ (implicit) & $0$ (implicit)\\
XSum \citep{narayan_xsum_2018} & Summarization & BBC source & $0$ (implicit) & $0$ (implicit)\\
\bottomrule
\end{tabularx}
\end{table*}

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{0.98}
\caption{\textbf{Factuality, faithfulness, and adjacent resources in the purposive mapping (part II).}
The coding follows Table~\ref{tab:full-mapping-factual-a}: $\kappa=0$ denotes factual only, and ``partial'' denotes incomplete representation of the content-permission policy.
These are author-coded profiles, not labels reported by the original resources.}
\label{tab:full-mapping-factual-b}
\begin{tabularx}{0.98\linewidth}{@{}P{0.23\textwidth}P{0.18\textwidth}P{0.20\textwidth}P{0.11\textwidth}Y@{}}
\toprule
\textbf{Resource} & \textbf{Task} & \textbf{$O$: reference evidence or rubric} & \textbf{Inferred $\sigma$} & \textbf{Inferred $\kappa$ / status}\\
\midrule
\multicolumn{5}{l}{\textbf{Strict factuality and faithfulness (continued)}}\\
CNN/DailyMail \citep{hermann_teaching_2015} & Summarization & News source & $0$ (implicit) & $0$ (implicit)\\
DelucionQA \citep{sadat_delucionqa_2023} & Domain QA & Manuals & $0$ (implicit) & $0$ (implicit)\\
HalluQA \citep{cheng_evaluating_2023} & Long-form QA & External factual reference & $0$ (implicit) & $0$ (implicit)\\
BIG-Bench \citep{srivastava_beyond_2023} & Multi-task reasoning & Task gold & $0$ (implicit) & $0$ (implicit)\\
BIG-Bench Hard \citep{suzgun_challenging_2023} & Reasoning & Task gold & $0$ (implicit) & $0$ (implicit)\\
CR-LLM \citep{li_cr-llm_2024} & Concept reasoning & Gold concepts/context & $0$ (implicit) & $0$ (implicit)\\
HotpotQA \citep{yang_hotpotqa_2018} & Multi-hop QA & Wikipedia & $0$ (implicit) & $0$ (implicit)\\
HypoTermQA \citep{uluoglakci_hypotermqa_2024} & Hypothetical QA & Task/world reference & $0$ (implicit) & $0$ (implicit)\\
HalluRAG \citep{ridder_hallurag_2025} & RAG QA & Retrieved Wikipedia & $0$ (implicit) & $0$ (implicit)\\
ToolQA \citep{zhuang_toolqa_2023} & Tool-use QA & Tools and APIs & $0$ (implicit) & $0$ (implicit)\\
RealTime QA \citep{kasai_realtime_2024} & Real-time QA & Current evidence & $0$ (implicit) & $0$ (implicit)\\
\midrule
\multicolumn{5}{l}{\textbf{Closest intent and hallucination taxonomies}}\\
FaithQA \citep{hao_faithqa_2025} & Intent verification & Query constraints & $0$ dominant; not controlled & $0$--$1$ (partial: intent constraints)\\
HIC-Bench \citep{yang_hicbench_2025} & Hallucination typing & Task-dependent & $0$ dominant; not controlled & $0$--$2$ (partial: hallucination types)\\
\bottomrule
\end{tabularx}
\end{table*}

\begin{table*}[t]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{0.98}
\caption{\textbf{Creativity and constrained problem-solving resources in the purposive mapping.}
Creative-writing resources tend toward $\sigma=1$--$2$ with task-implied $\kappa=2$, whereas divergent-thinking tasks often leave style uncontrolled.
Constrained problem-solving resources mostly use neutral form and partial $\kappa=1$ constraints.
The profiles are author annotations inferred from task design, not variables reported by the original resources.}
\label{tab:full-mapping-creative}
\begin{tabularx}{0.98\linewidth}{@{}P{0.23\textwidth}P{0.18\textwidth}P{0.20\textwidth}P{0.11\textwidth}Y@{}}
\toprule
\textbf{Resource} & \textbf{Task} & \textbf{$O$: reference evidence or rubric} & \textbf{Inferred $\sigma$} & \textbf{Inferred $\kappa$ / status}\\
\midrule
\multicolumn{5}{l}{\textbf{Creative generation and divergent thinking}}\\
CreativityPrism \citep{hou_creativityprism_2025} & Multi-task creativity & Task-specific rubrics & Mixed; not isolated & $2$ (implicit)\\
LitBench \citep{fein_litbench_2025} & Story evaluation & Human preferences & $1$--$2$ (implicit) & $2$ (implicit)\\
Short Story \citep{ismayilzada_evaluating_2025} & Story generation & Semantic/human criteria & $1$--$2$ (implicit) & $2$ (implicit)\\
WritingBench \citep{wu_writingbench_2025} & Creative writing & Rubric/human criteria & $1$--$2$ (style scored) & $2$ (implicit)\\
Art-or-Artifice / TTCW \citep{chakrabarty_art_2024} & Creative writing & Expert rubric & $1$--$2$ (implicit) & $2$ (implicit)\\
Automated DT scoring \citep{organisciak_beyond_2023} & Divergent thinking & Automated rubric & Not controlled & $2$ (implicit)\\
Verbal Creativity \citep{dinu_testing_2025} & Divergent thinking & TTCT & Not controlled & $2$ (implicit)\\
Automated Creativity \citep{li_automated_2025} & Creative writing & Reference texts & $1$--$2$ (implicit) & $2$ (implicit)\\
Hallucinating Narratives \citep{sun_hallucinating_2024} & Controlled story/reasoning & Mixed evidence/rubric & $1$--$2$ (implicit) & $2$ (partial; mixed support)\\
\midrule
\multicolumn{5}{l}{\textbf{Constrained creative problem solving}}\\
MacGyver \citep{tian_macgyver_2025} & Problem solving & Human feasibility & $0$ (implicit) & $1$ (partial: feasibility)\\
EscapeBench \citep{qian_escapebench_2025} & Agent puzzles & Environment state & $0$ (implicit) & $1$ (partial: correctness)\\
CreativEval \citep{delorenzo_creativeval_2024} & Code generation & Specification/tests & $0$ (implicit) & $1$ (partial: correctness)\\
NeoCoder \citep{lu_benchmarking_2025} & Code generation & Unit tests & $0$ (implicit) & $1$ (partial: correctness)\\
Math Creativity \citep{ye_assessing_2024} & Math problem solving & Human soundness & $0$ (implicit) & $1$ (partial: soundness)\\
\bottomrule
\end{tabularx}
\end{table*}

\clearpage
\bibliography{references}

\end{document}
