\documentclass[final]{clv2025}

\jvol{vv}
\jnum{nn}
\jyear{2025}
\dochead{Position Paper}

%% Choose your variant of English
\usepackage[american]{babel}

%% Required packages
\usepackage{natbib}
    \bibliographystyle{compling}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, calc}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{tabularx}
\usepackage{array}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{groupplots}



%% Drafting helpers (remove for camera-ready)
%% ------------------------------------------------------------
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{TODO:} #1}}
\newcommand{\citetodo}[1]{\textcolor{gray}{\textbf{[CITE:} #1\textbf{]}}}
\newcommand{\claimtodo}[1]{\textcolor{blue}{\textbf{[CLAIM:} #1\textbf{]}}}

%% ------------------------------------------------------------
%% Vocabulary (normalize throughout paper)
%% ------------------------------------------------------------
% \newcommand{\CI}{\textsc{CI}} % creative intent -- removed from the framework vocabulary
\newcommand{\TR}{\textsc{TR}} % truth regime
\newcommand{\TRs}{\textsc{TR}s} % truth regime (plural)
\newcommand{\PD}{\textsc{PD}} % productive divergence
\newcommand{\Hallu}{\textsc{H}} % hallucination (TR-violating)
\newcommand{\Faith}{\textsc{F}} % faithfulness
\newcommand{\Acc}{\textsc{A}} % accuracy/factuality

\title{Hallucination Is Relative: A Position on Truth-Regime-Aware Evaluation of LLM Divergence}

\runningtitle{Hallucination Is Relative}
\runningauthor{Carlesso et al.}

\author{Mathis Carlesso$^{1,2}$, Martino Lovisetto$^{2}$, Elöd Egyed-Zsigmond$^{1}$, Pierre-Yves Genest$^{2}$, Stefan Duffner$^{1}$}

\affilblock{
    \affil{INSA Lyon, CNRS, LIRIS, UMR5205, Université Claude Bernard Lyon 1, 69621 Villeurbanne, France\\\quad \email{mathis.carlesso@insa-lyon.fr}}
    \affil{Alteca, 69100 Villeurbanne, France}
}


\begin{document}
\maketitle

\begin{abstract}
Hallucination is usually scored as a binary defect: a statement is supported or it is not.
This misses a central property of language use: the same unsupported content can be dangerous in a medical answer, harmless in a metaphor, and useful in fiction or ideation.
We argue that hallucination should instead be evaluated relative to the prompt's \emph{truth regime}, $\TR(p)=(O,\sigma,\kappa)$: the oracle $O$ that grounds truth, the stylistic latitude $\sigma$ that permits surface variation, and the invention license $\kappa$ that determines whether new truth-conditional content may be introduced.
Under this view, hallucination (\Hallu) is a truth-regime violation, whereas productive divergence (\PD) is divergence that the regime licenses and the task rewards.
We then show that current hallucination and creativity benchmarks share a blind spot: neither conditions its scores on what the prompt authorizes.
We make this gap concrete without a new experiment, by auditing widely used benchmarks against $(O,\sigma,\kappa)$ and by applying a claim-level labeling rule to documented cases in which the same output flips between \Hallu\ and \PD\ as the regime changes.
These observations motivate a research agenda for truth-regime-aware evaluation that scores \Hallu\ and \PD\ separately under explicit oracle and permission annotations.
\end{abstract}


\section{Introduction: The Creativity--Hallucination Paradox}\label{sec:intro}

Large language models (LLMs) sometimes state things that are not true.
The community calls these errors \emph{hallucinations}: the model produces fluent, confident text that is not supported by evidence or contradicts known facts \citep{ji_survey_2023, huang_survey_2025}.
Hallucination is widely regarded as the central reliability barrier for deploying LLMs in high-stakes settings such as medicine, law, and education \citep{lin_truthfulqa_2022}.
A large body of work consequently aims to detect, measure, and reduce it \citep{li_halueval_2023, wei_measuring_2024, thorne_fever_2018}.

A paradox sits underneath this effort.
When users ask an LLM to write fiction, brainstorm, or propose creative solutions, they actively \emph{want} the model to go beyond established facts, and punishing every unsupported statement would suppress the very outputs they seek \citep{franceschelli_creativity_2024, hou_creativityprism_2025, rick_supermind_2023}.
A brainstorming tool that only repeats ideas that already exist is useless; a medical assistant that invents diagnoses is dangerous.
The difference lies not in the text itself, but in what the prompt asked for.

\paragraph{The diagnostic gap.}
Current evaluation frameworks do not register this difference.
Hallucination benchmarks treat any unsupported content as a defect to minimize \citep{bang_hallulens_2025}, whereas creativity benchmarks treat novelty as a virtue to maximize, often without checking factual grounding \citep{hou_creativityprism_2025, acar_creativity_2019}.
Neither tradition conditions its scores on what the user's prompt actually authorizes.
As a result, one output can be scored as a serious failure by one framework and a creative success by another, depending on which protocol is applied rather than on whether the divergence was appropriate.

Even the closest recent steps stop short of the contract. Intent-aware evaluation decomposes a query into checkable constraints, but those constraints cover query elements (subject, action, quantity) rather than stylistic latitude or invention license, and it scores intent coverage with no truth oracle: it has neither $O$ nor $\kappa$ \citep{hao_faithqa_2025}. The closest analogue to our distinction splits hallucinations into ``intelligent'' and ``defective,'' yet it reads that label off the output's own creativity and factual-deviation scores rather than off the prompt, so the label is fixed by the text and cannot flip with the contract that should govern it \citep{yang_hicbench_2025}. Neither makes $(O,\sigma,\kappa)$ a scored variable, and that omission is the gap we identify.

This gap cannot be closed simply by building better detectors.
Theoretical work shows that LLMs cannot avoid hallucination entirely: arguments from incompleteness and undecidability \citep{banerjee_llms_2024, xu_hallucination_2024} and from calibration \citep{kalai_why_nodate} establish a nonzero structural error floor.
Yet the same generative mechanism also produces the novel hypotheses and creative combinations that users request in open-ended tasks \citep{sun_hallucinating_2024, he_shakespearean_2025}.
The goal, therefore, cannot be to eliminate divergence, but to judge it against what the task permits.

\paragraph{Our thesis.}
Hallucination is not a binary property of text.
It is \emph{relative to the prompt's truth regime} (\TR): the task-specific contract $\TR(p)=(O,\sigma,\kappa)$ specifying the oracle $O$ against which claims are checked, the stylistic latitude $\sigma$ granted to the form of the output, and the invention license $\kappa$ that determines whether new truth-conditional content may be introduced.
Under a strict \TR\ (e.g., ``summarize this document accurately''), any unsupported claim is a hallucination (\Hallu).
Under a permissive \TR\ (e.g., ``write fiction''), that same divergence may instead constitute \emph{productive divergence} (\PD): a regime-compliant departure from established facts that is novel and task-useful, and therefore fulfills rather than violates what the prompt requested.
The distinction turns not on the text but on whether the prompt authorized the divergence; evaluation that ignores this conflates \Hallu\ with \PD.
Concretely, the distinction is already actionable: an evaluator who annotates a prompt's oracle and invention license can withhold the hallucination penalty wherever the regime authorized the divergence, scoring it as \PD\ rather than as error, a judgment no regime-agnostic scorer can make.
This is also what separates \TR\ from context-faithfulness: faithfulness asks only whether an output stays within its evidence, and has no analogue of $\kappa$, the explicit license to leave it.

\paragraph{How we establish the gap.}
We argue this position rather than test it with a new benchmark.
Two analytical moves carry the weight.
First, we audit widely used hallucination and creativity benchmarks against the three components $(O,\sigma,\kappa)$ and find that, while the oracle is sometimes fixed, stylistic latitude and invention license are essentially never annotated, so the contract remains implicit (Section~\ref{sec:sota}).
Second, we apply a claim-level labeling rule to documented cases and show that the same output receives opposite verdicts (\Hallu\ under one regime, \PD\ under another) once the contract is made explicit (Section~\ref{sec:worked}).
Together these establish that the gap is structural, not anecdotal, and that a minimal evaluation layer making $(O,\sigma,\kappa)$ explicit would resolve verdicts that current pipelines leave underdetermined.

\paragraph{Contributions.}
This position paper makes three contributions.
\emph{First} (conceptual), we introduce $\TR(p)=(O,\sigma,\kappa)$ and a minimal vocabulary (\TR, \PD, \Hallu) that makes explicit what the prompt authorizes, separating the two components that bear on truth (the oracle $O$ and the invention license $\kappa$) from the stylistic latitude $\sigma$ that governs form and conditions measurement validity (Section~\ref{sec:framework}).
\emph{Second} (diagnostic), we show that hallucination and creativity evaluation share a blind spot (neither conditions its scores on the truth regime), and we demonstrate it concretely by auditing existing benchmarks and by applying the labeling rule to documented cases (Sections~\ref{sec:sota}~and~\ref{sec:worked}).
\emph{Third} (programmatic), we outline a research agenda for \TR-aware benchmarks that score \Hallu\ and \PD\ separately under explicit oracle and permission annotations (Section~\ref{sec:agenda}).


\section{Terminology and Conceptual Framework}\label{sec:framework}

\paragraph{Motivation.}
The literature uses the word \emph{hallucination} broadly, but not consistently.
Surveys commonly group together factuality errors against world knowledge and faithfulness errors against a provided source, sometimes adding axes such as intrinsic vs.\ extrinsic causes and different granularities \citep{ji_survey_2023, huang_survey_2025, zhang_sirens_nodate}.
Benchmarking work further argues that hallucination is often conflated with factuality and that any verdict depends on the chosen oracle and on evaluator reliability \citep{bang_hallulens_2025}.
To reason about creativity--hallucination trade-offs, we fix a minimal vocabulary that makes the prompt's \emph{truth regime} explicit.
Throughout, we separate \emph{prompt-level} variables (what the user requests) from \emph{output-level} properties (what the model produces), and adopt a \emph{claim-level} unit of analysis because evaluations vary by granularity.

\textbf{Truth regime (\TR; prompt-level contract).}
A task-specific contract $\TR(p)=(O,\sigma,\kappa)$ that determines what must be \emph{grounded}, how freely the \emph{form} may vary, and what may be \emph{invented}.
\TR\ is needed because both the definition of ``hallucination'' and the evaluation oracle shift across settings \citep{bang_hallulens_2025}.
We treat $\TR(p)$ as an \emph{idealization}: in deployment the contract is shaped not only by the visible prompt but also by system instructions, domain conventions, and user expectations, and an ambiguous prompt may underdetermine it.
Inferring \TR\ under such ambiguity is itself an open problem (Section~\ref{sec:agenda}); we nonetheless locate the contract at the prompt because that is where evaluation can render it explicit and auditable.

\textbf{Oracle ($O$).}
The reference against which truth-conditional claims are checked: a source document, a retrieval set, a database, world knowledge, or user-provided facts.
The same claim can be supported by one oracle and unsupported by another, so $O$ must be specified rather than assumed \citep{bang_hallulens_2025, ji_survey_2023}.

\textbf{Stylistic latitude ($\sigma$).}
How freely the \emph{form} of the output may vary while preserving its propositions.
This axis is studied independently of content: style is not a single variable but a bundle of co-varying dimensions \citep{kang_style_2021}, it can be transferred while meaning is held fixed \citep{jin_deep_2022}, and the dual operation (moving facts across a fixed style) is equally well defined \citep{balepur_text_2023}.
Figurative devices such as metaphor and hyperbole are formal deviations that leave truth conditions intact \citep{badathala_match_2023}, and foregrounding, the long-standing notion that stylistically marked language stands out against an unmarked background \citep{leech_short_style_1981}, is reliably recognized by readers \citep{vanpeer_stylistics_2020}, which is what lets a graded $\sigma$ be annotated rather than guessed; established formality datasets and style-transfer metrics make the axis operationally measurable \citep{pavlick_empirical_2016, rao_dear_2018, briakou_evaluating_2021}.
Three illustrative levels suffice: $\sigma{=}0$, literal and neutral form (no marked figures or authorial voice); $\sigma{=}1$, moderately marked form (didactic analogies, occasional address to the reader); $\sigma{=}2$, strongly marked form (sustained imagery, salient voice).
Three levels, rather than two or five, keep an explicit median anchor (the contested ``moderately marked'' zone a binary split would erase) while omitting the extreme registers that rarely surface in model output.
By construction, $\sigma$ licenses variation in form only: a metaphor that preserves the underlying proposition introduces no new truth-conditional content.
Two consequences fix $\sigma$'s role in the framework.
First, $\sigma$ is orthogonal to truth: since it neither adds nor removes a proposition, it cannot by itself make a claim \Hallu\ or \PD; those verdicts turn only on $O$ and $\kappa$, the two components that bear on content.
Second, $\sigma$ is nonetheless load-bearing for \emph{measurement}: raising stylistic latitude is exactly what leads an automatic verifier to misread marked form as unsupported content, so a regime that leaves $\sigma$ unspecified leaves the validity of any \Hallu\ count undefined.
$\sigma$ thus enters the contract not as a third verdict axis but as the dimension along which measurement of the other two must be held invariant (Section~\ref{sec:agenda}).

\textbf{Invention license ($\kappa$).}
Whether new truth-conditional content may be introduced.
Again, three illustrative levels: $\kappa{=}0$, no invention (every truth-conditional claim must be supported by $O$); $\kappa{=}1$, flagged speculation (hypotheses and extrapolations are permitted provided they are marked as such); $\kappa{=}2$, free invention within a declared frame (fiction, ideation), bound by coherence and user-provided facts rather than by $O$.
Theoretically, $\kappa$ tracks the kind of speech act the prompt licenses: at $\kappa{=}0$ the model must make genuine assertions, committing to the truth of what it adds to the common ground \citep{stalnaker_assertion_1978}, whereas a declared fictional frame ($\kappa{=}2$) licenses what \citet{searle_logical_1975} calls pretended assertions, utterances with the form of assertion but without the commitment to truth, with $\kappa{=}1$ in between, admitting assertions explicitly marked as provisional.
The two permissions are independent: many real prompts request creative style without factual invention (e.g., ``explain photosynthesis like a poet''), so $\sigma$ is high while $\kappa$ stays at zero, and figurative language is welcome but fabricated claims are not \citep{franceschelli_creativity_2024}.
Throughout, we write $\sigma$ or $\kappa$ ``low''/``high'' for the bottom/top of these scales.

\textbf{Faithfulness (\Faith; output-level).}
Whether the output is \emph{supported by the provided evidence} (documents, retrieved passages, tables), without introducing unsupported details.
This is the central failure mode in grounded generation and summarization, where fluent outputs can exceed source support \citep{maynez_faithfulness_2020, kryscinski_evaluating_2020}.

\textbf{Accuracy / factuality (\Acc; output-level).}
Whether the claims that must be true under \TR\ are actually supported by an appropriate external source (e.g., references, gold labels, or world knowledge).
This corresponds to question answering (QA) and fact-checking evaluation.

\textbf{Hallucination (\Hallu; claim-level).}
As we define it, a hallucination is a \TR-violation: the model presents a claim \emph{as grounded/true} where \TR\ requires truth or evidence, but the claim is unsupported (\Faith) and/or incorrect (\Acc).
This unifies faithfulness and factuality failures while making the oracle explicit \citep{maynez_faithfulness_2020, bang_hallulens_2025}.

\textbf{Productive divergence (\PD; claim-level).}
Content that would count as hallucination under a strict factual standard, but is allowed by the prompt's truth regime and helps accomplish the task.
A claim is \PD\ when three conditions hold: it is \emph{licensed} (the regime permits invention, $\kappa>0$, or the claim falls inside a declared fictional frame); it is \emph{useful} (novel and appropriate under the task constraints, the standard creativity criterion \citep{diedrich_are_2015, acar_creativity_2019}); and it is \emph{presented on the regime's terms} (a flagged hypothesis under $\kappa{=}1$ must not be asserted as verified fact).
This echoes non-LLM uses of ``productive divergence'' as variation that energizes performance \citep{parry_team_2023, lingard_pulling_2017}.
A contrast makes the category concrete.
Given ``summarize this medical report,'' an added claim that the patient has diabetes, absent from the report, violates a regime where $O$ is the report and $\kappa{=}0$: it is \Hallu.
Given ``write a poem about quantum physics,'' the line that electrons dream in parallel worlds is supported by no oracle, yet the regime licenses figurative invention: it is \PD\ if it serves the poem.
Both statements are invented; only the first is a failure.
Crucially, if \TR\ demands grounding, the same divergence is \Hallu, not \PD.

\noindent\textbf{Decision rule.}
Given a prompt $p$, an output $y$, and an oracle $O$: specify $\TR(p)$, decompose $y$ into atomic claims, then label each claim using $O$ as \Hallu\ (truth-conditional under $\TR$ but unsupported/incorrect), \PD\ (licensed by $\TR$ and task-useful), or neither (stylistic variation, benign noise); Appendix~\ref{app:decision-rule} gives the step-by-step procedure.
The support condition is mechanically checkable against $O$; the usefulness condition inherits the evaluator dependence of creativity judgments (Section~\ref{sec:agenda}).

\paragraph{Worked example.}
Consider ``Explain photosynthesis in a poetic style, but keep the science accurate.''
Here the oracle is scientific world knowledge, $\sigma{=}2$ because figurative language is requested, and $\kappa{=}0$ because the science must stay accurate.
Calling chlorophyll ``a green workshop for sunlight'' is stylistic variation, but claiming that plants make sugar from moonlight is \Hallu.
If the prompt instead asks for a fictional plant ecology where moonlight powers growth, the same moonlight claim can be \PD: useful, clearly framed as invented, and consistent with the fictional constraints.

\noindent Under this vocabulary we do not oppose ``creativity vs.\ truth.''
We oppose \textbf{\Hallu} (\TR\ violations) to \textbf{\PD} (\TR-compliant creativity); regime-agnostic factuality evaluation fails to separate the two.

% ============================================================
% Truth regimes + divergences (conceptual table, with concrete examples)
% ============================================================
\begin{table*}[tb]
\centering
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.3}
\caption{\textbf{Canonical truth regimes with concrete examples.}
$\sigma$ = stylistic latitude; $\kappa$ = invention license.
The verdict on the \emph{same} surface content turns on $(O,\kappa)$, not on the text alone:
a claim that is \Hallu\ under a strict regime can be \PD\ under a permissive one
\citep{maynez_faithfulness_2020,kryscinski_evaluating_2020,franceschelli_creativity_2024}.}
\label{tab:canonical-truth-regimes}
\begin{tabularx}{\linewidth}{%
  >{\bfseries}p{1.9cm}                      % Regime
  p{2.2cm}                                   % Oracle
  >{\centering\arraybackslash}p{1.55cm}      % (sigma,kappa)
  p{2.7cm}                                   % Prompt example
  X                                          % Example PD
  X                                          % Example H
}
\toprule
\textbf{Regime} & \textbf{Oracle $O$} & \textbf{$(\sigma,\kappa)$} &
\textbf{Prompt example} &
\textbf{Example \PD\ claim} &
\textbf{Example \Hallu\ claim} \\
\midrule
Strict grounded generation / summarization &
Provided sources (docs / retrieved passages) &
(low, 0) &
\textit{``Summarize this medical report accurately.''} &
--- \emph{(regime admits none)} &
``The patient has Type 2 diabetes'' when absent from the report. \\
\midrule
Factual task with creative style &
World knowledge and/or cited sources &
(high, 0) &
\textit{``Explain photosynthesis in a poetic style, but keep the science right.''} &
``Chlorophyll---a green workshop for sunlight'' \emph{(figurative; proposition intact)}. &
``Plants make sugar from moonlight'' \emph{(factually incorrect claim introduced)}. \\
\midrule
Ideation / fiction-writing &
Prompt constraints (coherence, user-provided facts / outline) &
(high, 2) &
\textit{``Write a sci-fi story where moonlight powers plant growth.''} &
``Photovores harvest lunar energy via chloromoon'' \emph{(invented; coherent; on-brief)}. &
``Photovores run on electricity'' \emph{(contradicts the user-specified constraint)}. \\
\bottomrule
\end{tabularx}
\end{table*}

\subsection{Truth regimes and regime-relative divergence}\label{sec:truth-regimes}

The components of $\TR(p)=(O,\sigma,\kappa)$ combine into recognizable task profiles \citep{ji_survey_2023,huang_survey_2025}.
Table~\ref{tab:canonical-truth-regimes} instantiates three canonical regimes.
Existing benchmarks typically fix $O$ but rarely annotate $(\sigma,\kappa)$ explicitly, making it easy to conflate \PD\ with \Hallu\ \citep{bang_hallulens_2025}.

\paragraph{Claim-level labeling.}
Given a regime $\TR(p)$, each atomic claim falls into one of the categories already defined: \emph{stylistic variation} (form changes, no new propositions), \PD\ (novel content licensed by $\kappa$ and task-useful), \Hallu\ (truth-conditional content unsupported by oracle $O$), or \emph{benign noise} (minor inaccuracies with negligible impact) \citep{ji_survey_2023,bang_hallulens_2025}.
These four categories are the leaves of the decision rule's three branches: \Hallu\ and \PD\ are the two scored verdicts, while stylistic variation and benign noise jointly form its \emph{neither} branch.

\noindent\textbf{Two-sided failure relative to $\kappa$.}
When $\kappa$ is low, excess divergence increases \Hallu; when $\kappa$ is high, \emph{insufficient} divergence yields conservative outputs that fail the creative objective, a failure mode captured by creativity benchmarks but invisible to hallucination-only scoring \citep{hou_creativityprism_2025}.
The two sides sit at different levels: excess divergence is visible claim by claim, whereas insufficient divergence is an output-level absence, a distinction we make precise when defining verdicts (Section~\ref{sec:discussion}).
A regime-aware evaluator must therefore be able to fail an output in either direction, which a single unsupportedness count cannot.

With this vocabulary and labeling procedure in place, the natural question is whether existing evaluation frameworks already condition their scores on the truth regime.
The next section examines both hallucination and creativity benchmarks and shows that they do not.

\section{The Diagnostic Gap: Evaluation That Ignores the Truth Regime}\label{sec:sota}

\paragraph{Two traditions, one shared blind spot.}
Hallucination benchmarks treat divergence from an oracle as a defect to minimize \citep{ji_survey_2023, bang_hallulens_2025}; creativity benchmarks treat it as evidence of novelty to maximize \citep{hou_creativityprism_2025}.
The split works for single-intent tasks but breaks down for mixed-intent prompts that pair novelty with factual constraints \citep{wang_user-centric_2024}: the same surface divergence can be \PD\ or \Hallu, yet on both sides the truth regime $(O,\sigma,\kappa)$ stays implicit.
Table~\ref{tab:audit} makes the omission concrete.
Auditing forty benchmarks, organized by \emph{what} they evaluate (hallucination/faithfulness vs.\ creativity/divergence) and by the \emph{task} they score, we find the same pattern throughout: the oracle $O$ is frequently fixed, but stylistic latitude $\sigma$ is never annotated and the invention license $\kappa$ is never graded.
Hallucination suites inherit a fixed $O$ and assume $\kappa$ low; creativity suites drop the factual oracle and leave $\kappa$ high but unscored; the few intent-aware benchmarks condition on the query without separating $\sigma$ from $\kappa$.
The contract is left to the task label.

\paragraph{The uniform columns are the finding, not an empty table.}
One might object that columns repeating the same entry across forty benchmarks carry little information.
The opposite holds: the uniformity \emph{is} the diagnostic result.
Were stylistic latitude or invention license sometimes annotated, the columns would vary and the blind spot would be partial and design-specific; that every benchmark, across both research traditions and every task type, leaves $\sigma$ unannotated and $\kappa$ ungraded is what shows the omission to be structural rather than incidental.
The two columns also fail in different ways, and the table records which: $\sigma$ is absent everywhere, whereas $\kappa$ is never missing at random but pinned to an assumed value (low under hallucination suites, high but unscored under creativity suites), so the audit reveals not only that the contract goes unannotated but \emph{how} each tradition silently fixes it.

\paragraph{How the audit was coded.}
We count a benchmark as annotating $\sigma$ only when stylistic latitude is an explicit, scored variable in its protocol, not merely present in some prompts, and as grading $\kappa$ only when invention is a scored dimension with more than one licensed level.
By this criterion a creative-writing suite that varies the style of its prompts without making latitude a scored variable still counts as leaving $\sigma$ unannotated, since the latitude never enters the score.
The coding is a single-author pass against these criteria, offered as the first instance of the annotation layer we advocate rather than as a validated resource; because the test is near-binary (is $(\sigma,\kappa)$ a scored variable?), it leaves little room for coder disagreement, and independent re-coding with reported agreement is itself part of the agenda (Section~\ref{sec:agenda}).

\begin{table*}[!tp]
\centering
\scriptsize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.1}
\caption{\textbf{Audit of forty benchmarks against the truth-regime components $(O,\sigma,\kappa)$, grouped by what they evaluate and by task.}
$\sigma$ asks whether stylistic latitude is an annotated, scored variable; $\kappa$ likewise for the invention license.
Legend: ``--'' = not represented; ``0'' = invention not licensed (assumed by the task, not annotated); ``hi'' = invention is central but left unscored.
No benchmark annotates $\sigma$, and none grades $\kappa$; oracle $O$ is the only component routinely fixed.
TTCT/TTCW: Torrance Tests of Creative Thinking/Writing; RAG: retrieval-augmented generation.}
\label{tab:audit}
\begin{tabular}{p{0.335\textwidth} p{0.135\textwidth} p{0.165\textwidth} c c p{0.155\textwidth}}
\toprule
\textbf{Benchmark} & \textbf{Task evaluated} & \textbf{Oracle $O$} & \boldmath$\sigma$? & \boldmath$\kappa$? & \textbf{Regime} \\
\midrule
\addlinespace[1pt]
\multicolumn{6}{l}{\textbf{Evaluates hallucination / faithfulness}} \\
\addlinespace[1pt]
FEVER \citep{thorne_fever_2018} & Fact-checking & Wikipedia evidence & -- & 0 & Implicit (strict) \\
TruthfulQA \citep{lin_truthfulqa_2022} & Factual QA & World knowledge & -- & 0 & Implicit (strict) \\
PopQA \citep{mallen_when_2023} & Long-tail QA & Wikidata & -- & 0 & Implicit (strict) \\
SimpleQA \citep{wei_measuring_2024} & Factual QA & World knowledge & -- & 0 & Implicit (strict) \\
TriviaQA \citep{joshi_triviaqa_2017} & Reading-comp.\ QA & Evidence documents & -- & 0 & Implicit (strict) \\
DefAn \citep{rahman_defan_2024} & Factual QA & World knowledge & -- & 0 & Implicit (strict) \\
HaluEval \citep{li_halueval_2023} & QA + summarization & World knowl.\ + ctx & -- & 0 & Implicit (strict) \\
FactBench \citep{bayat_factbench_2024} & Factual QA & World knowledge & -- & 0 & Implicit; \texttt{undecidable} \\
HalluLens \citep{bang_hallulens_2025} & Factual QA & World knowledge & -- & 0 & Implicit (strict) \\
FRANK \citep{pagnoni_understanding_2021} & Summarization & Source (news) & -- & 0 & Implicit (strict) \\
QAGS \citep{wang_asking_2020} & Summarization & Source (news) & -- & 0 & Implicit (strict) \\
SummEval \citep{fabbri_summeval_2021} & Summarization & Source (news) & -- & 0 & Implicit (strict) \\
XSum \citep{narayan_xsum_2018} & Summarization & Source (BBC) & -- & 0 & Implicit (strict) \\
CNN/DailyMail \citep{hermann_teaching_2015} & Summarization & Source (news) & -- & 0 & Implicit (strict) \\
DelusionQA \citep{sadat_delucionqa_2023} & Domain QA & Source (manuals) & -- & 0 & Implicit (strict) \\
HaluQA \citep{cheng_evaluating_2023} & Long-form QA & World knowledge & -- & 0 & Implicit (strict) \\
BIG-Bench \citep{srivastava_beyond_2023} & Multi-task reasoning & Task gold & -- & 0 & Implicit (strict) \\
BIG-Bench Hard \citep{suzgun_challenging_2023} & Reasoning & Task gold & -- & 0 & Implicit (strict) \\
CR-LLM \citep{li_cr-llm_2024} & Concept reasoning & Concept gold + ctx & -- & 0 & Implicit (strict) \\
HotpotQA \citep{yang_hotpotqa_2018} & Multi-hop QA & Wikipedia & -- & 0 & Implicit (strict) \\
HypoTermQA \citep{uluoglakci_hypotermqa_2024} & Hypothetical QA & World knowledge & -- & 0 & Implicit (strict) \\
HalluRAG \citep{ridder_hallurag_2025} & RAG QA & Retrieved Wikipedia & -- & 0 & Implicit (strict) \\
ToolQA \citep{zhuang_toolqa_2023} & Tool-use QA & Tools / APIs & -- & 0 & Implicit (strict) \\
RealTimeQA \citep{kasai_realtime_2024} & Real-time QA & Current events & -- & 0 & Implicit (strict) \\
FaithQA \citep{hao_faithqa_2025} & Intent verification & Query constraints & -- & -- & Partial; intent only \\
HIC-Bench \citep{yang_hicbench_2025} & Hallucination typing & Task-dependent & -- & -- & Partial; intel./defect. \\
\midrule
\addlinespace[2pt]
\multicolumn{6}{l}{\textbf{Evaluates creativity / divergence}} \\
\addlinespace[1pt]
CreativityPrism \citep{hou_creativityprism_2025} & Multi-task creativity & None (rubric) & -- & hi & Absent \\
LitBench \citep{fein_litbench_2025} & Story evaluation & None (human pref.) & -- & hi & Absent \\
Short Story \citep{ismayilzada_evaluating_2025} & Story generation & None (sem./human) & -- & hi & Absent \\
WritingBench \citep{wu_writingbench_2024} & Creative writing & None (rubric/human) & -- & hi & Absent \\
Art-or-Artifice / TTCW \citep{chakrabarty_art_2024} & Creative writing & None (expert rubric) & -- & hi & Absent \\
SemDis \citep{organisciak_beyond_2023} & Divergent thinking & None (embed.\ dist.) & -- & hi & Absent \\
Verbal Creativity \citep{dinu_testing_2025} & Divergent thinking & None (TTCT) & -- & hi & Absent \\
Automated Creativity \citep{li_automated_2025} & Creative writing & Reference texts & -- & hi & Absent \\
Hallu.\ Narratives \citep{sun_hallucinating_2024} & Controlled story & Mixed & -- & hi & Absent; unlabeled \\
MacGyver \citep{tian_macgyver_2025} & Problem-solving & Feasibility (human) & -- & hi & Partial; correctness \\
EscapeBench \citep{qian_escapebench_2025} & Agent puzzles & Env.\ state (auto) & -- & hi & Partial; correctness \\
CreativEval \citep{delorenzo_creativeval_2024} & Code generation & Spec / correctness & -- & hi & Partial; correctness \\
NeoCoder \citep{lu_benchmarking_2025} & Code generation & Unit tests & -- & hi & Partial; correctness \\
Math Creativity \citep{ye_assessing_2024} & Math problem-solving & Soundness (human) & -- & hi & Partial; correctness \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Hallucination evaluation assumes a strict regime}
Hallucination evaluation is anchored in the choice of an oracle, either \emph{world knowledge} (\Acc: factuality) or \emph{provided context} (\Faith: faithfulness), with further axes for contradiction vs.\ fabrication and evaluation granularity \citep{ji_survey_2023, zhang_sirens_nodate, bang_hallulens_2025}.
Modern protocols rely on claim-level decomposition and automated verification, e.g., retrieval with natural language inference (NLI) \citep{chen_menli_2023} or QA-based checking \citep{wang_asking_2020}, building on fact-checking \citep{thorne_fever_2018}.
These pipelines are technically mature: atomic-claim decomposition is established \citep{min_factscore_2023}, joint decomposition and verification run at scale \citep{rajendhran_verifastscore_2025}, some verifiers add a third \texttt{undecidable} verdict \citep{bayat_factbench_2024}, and complementary signals such as semantic entropy \citep{farquhar_detecting_2024} or internal probes \citep{marks_geometry_2023} flag unsupported content without consulting any prompt-level contract.
Yet these axes are necessary, not sufficient, for \TR-aware judgment: most pipelines assume a strict invention license ($\kappa$ low), so any unsupported truth-conditional content is a failure.
The assumption is defensible in grounded generation but collapses when prompts license speculation, and even a three-way verdict with \texttt{undecidable} \citep{bayat_factbench_2024} refines the label space without consulting the contract: licensed invention is not merely undecidable, it is \emph{authorized}.

\subsection{Creativity evaluation leaves the oracle implicit}
Creativity is commonly defined as combining \emph{novelty} with \emph{value/appropriateness} under constraints \citep{runco_standard_2012}, but operationalizations vary sharply across tasks \citep{acar_creativity_2019, franceschelli_creativity_2024}.
Recent LLM work treats creativity as multi-dimensional, with benchmarks spanning divergent-thinking rubrics, creative-writing evaluation, and constrained problem solving; \textsc{CreativityPrism} typifies the trend by aligning heterogeneous tasks to shared quality, novelty, and diversity dimensions \citep{hou_creativityprism_2025, fein_litbench_2025}.
Such scores also remain evaluator-dependent: ratings shift with rubric design and rater background, changing model rankings even under identical prompts when using LLM-as-judge protocols \citep{ismayilzada_evaluating_2025, liu_g-eval_2023, zheng_judging_2023}.
From a \TR\ perspective, the deeper issue is that creativity benchmarks rarely treat the truth regime as an explicit variable: novelty can be rewarded without distinguishing \PD\ from \Hallu\ \citep{sui_confabulation_2024, padmakumar_measuring_2025}, even when factual grounding is part of the contract \citep{wang_user-centric_2024}.
Faithfulness metrics could detect such violations \citep{maynez_faithfulness_2020, chen_menli_2023}, but they are not integrated into creativity pipelines.

\subsection{The shared failure mode and the closest prior work}
When $\TR$ is left implicit, two symmetric errors arise.
Under a permissive regime ($\kappa$ high), strict factuality checks misclassify licensed speculation as \Hallu\ (false positives on \PD) \citep{cyriac_metrics_2025, sui_confabulation_2024}; under a strict regime ($\kappa$ low), novelty-oriented scores reward confident fabrications that inflate apparent originality (false negatives on \Hallu) \citep{liu_g-eval_2023, chen_menli_2023}.
The closest existing steps narrow but do not close the gap.
Intent-hallucination evaluation decomposes a query into intent constraints (subject, action, qualifiers, quantity, and the like) and conditions the verdict on whether the response covers them, but it defines failure independently of factual accuracy, supplies no oracle, and treats even flagged invention as a misinterpretation rather than a licensed move: it encodes neither $O$ nor $\kappa$ \citep{hao_faithqa_2025}.
A recent benchmark splits hallucinations into ``intelligent'' and ``defective,'' the closest analogue to the \PD/\Hallu\ distinction, but it reads that label off the output's creativity and factual-deviation scores rather than off the prompt, so the label is determined by the text and cannot flip with the contract that should govern it \citep{yang_hicbench_2025}.
Neither makes $(O,\sigma,\kappa)$ a scored variable; that minimal layer is what we argue for.

\subsection{Why universal suppression of \Hallu\ is neither feasible nor desirable}
LLMs are probabilistic next-token predictors trained under incomplete supervision; they emit plausible continuations even when truth is underspecified, so eliminating \Hallu\ universally would require restricting the model or the task class.
Formal arguments support this floor \citep{banerjee_llms_2024, xu_hallucination_2024}, and strict regimes still reveal large residual error rates across short-form QA and truthfulness benchmarks \citep{mallen_when_2023, lin_truthfulqa_2022}.
Alignment tuning compounds the picture rather than resolving it: policies for refusal, hedging, and abstention learned during alignment tuning are themselves implicit, provider-chosen regime commitments that reshape when a model declines to diverge \citep{ouyang_training_2022}.
Crucially, the same generative mechanism that produces unsupported claims also produces the novel hypotheses users request: \citet{sun_hallucinating_2024} formalize ``good hallucinations'' as reasoning paths that remain correct after convergence, and \citet{banerjee_does_2025} show that verification-style mitigation can preserve divergent thinking whereas some decoding interventions suppress it.
Because \Hallu\ has a structural floor and the same divergence drives \PD, the practical goal is not universal suppression but \TR-calibrated control: specify the regime, license divergence where the regime allows it, and enforce convergence before anything is presented as fact \citep{bang_hallulens_2025, he_shakespearean_2025}.

\section{Applying the Framework: Worked Verdicts}\label{sec:worked}

The audit shows that the contract is missing; this section shows that supplying it \emph{changes verdicts}.
We apply the decision rule of Section~\ref{sec:framework} by hand to three illustrative cases.
Each case is drawn from a documented phenomenon: human-annotated XSum hallucinations (A), fictional role-play hallucination (B), and FactBench's \texttt{undecidable} label (C). Each isolates a verdict that current pipelines leave underdetermined, and in each the label flips between \Hallu\ and \PD\ when only $(O,\sigma,\kappa)$ changes.
Table~\ref{tab:worked} summarizes the flips.

\paragraph{Case A: correct but unsupported (oracle shift).}
\citet{maynez_faithfulness_2020} document this pattern directly on XSum, where annotators label a summary span as an \emph{extrinsic but factual} hallucination: content not found in the source document yet factually correct.
Their worked instance is a system summary that supplies the correct year of a London mayoral election (``2016''), accurate against world knowledge but absent from the article.
Decomposing the output to the claim level isolates this token from the system's other, incorrect additions.
Under the summarization regime $O=$ the source, $\sigma$ low, $\kappa{=}0$, the year is truth-conditional and unsupported by $O$: it is \Hallu, even though it satisfies factuality (\Acc) against world knowledge.
Hold the output fixed and change only the oracle (``Tell me about this election,'' with $O=$ world knowledge), and the same token becomes supported, hence not \Hallu.
The flip is exactly the unresolved tension Maynez et al.\ flag, whether summaries should tolerate such factual hallucinations \citep{maynez_faithfulness_2020, kryscinski_evaluating_2020}; the framework attributes it to $O$ rather than to the text.

\paragraph{Case B: novelty with an embedded factual commitment ($\kappa$ split within one output).}
Fictional role-play makes this split routine: \citet{sadeq_mitigating_2024} document that characters in a declared fictional frame act out of character and hallucinate about matters outside the scope of their knowledge, inserting real-world claims the frame does not support.
The pattern instantiates cleanly. Asked to ``write a short story set in 1920s Paris,'' a model invents a café and its proprietor (licensed under $\kappa{=}2$ within the fictional frame, hence \PD\ if they serve the story), but also has a real, named historical figure assert a specific claim that figure never made, presented as biographical fact.
The two are different speech acts: the invented café is a pretended assertion the frame authorizes \citep{searle_logical_1975}, whereas the biographical claim is a genuine assertion about the real world and so steps outside the frame.
It therefore fails the \emph{presented-on-the-regime's-terms} condition and is \Hallu, not \PD.
A novelty-oriented score would reward both as inventive \citep{hou_creativityprism_2025, sui_confabulation_2024}; the framework separates the licensed invention from the truth-commitment violation inside the same text.

\paragraph{Case C: undecidable is not authorized (license shift).}
FactBench's VERIFY pipeline labels each verifiable unit \texttt{supported}, \texttt{unsupported}, or \texttt{undecidable}, the last when retrieved evidence is insufficient to confirm or refute it, and its hallucination score penalizes \texttt{undecidable} units at a fixed weight regardless of what the prompt asked for \citep{bayat_factbench_2024}.
A documented instance is a model claim that a tool ``supports multiple input formats,'' which VERIFY marks \texttt{undecidable} because support varies across versions.
That label tracks evidence availability, not the regime: under $\kappa{=}0$, asserting such a claim as established fact is \Hallu\ and the penalty is warranted; under $\kappa{=}1$ (flagged speculation), the same epistemically undecidable content, marked as a hypothesis, is \PD.
The epistemic status (undecidable) is constant; the verdict is set by the license $\kappa$, which a fixed undecidable penalty, and even a three-way verdict, cannot see.

\begin{table}[htbp]
\centering
\small
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.2}
\caption{Worked verdicts: the same claim flips between \Hallu\ and \PD\ when one component of $(O,\sigma,\kappa)$ changes.}
\label{tab:worked}
\begin{tabular}{@{}p{0.05\columnwidth} p{0.33\columnwidth} p{0.27\columnwidth} p{0.27\columnwidth}@{}}
\toprule
\textbf{Case} & \textbf{Claim} & \textbf{Regime that yields \Hallu} & \textbf{Regime that yields \PD/ok} \\
\midrule
A & correct event year (absent from source) & $O=$ source, $\kappa{=}0$ & $O=$ world knowledge \\
B & real figure asserts words they never said & fictional frame, but stated as fact & ($\kappa{=}2$ applies to the invented café, not this claim) \\
C & undecidable claim (insufficient evidence) & $\kappa{=}0$, asserted as fact & $\kappa{=}1$, flagged as hypothesis \\
\bottomrule
\end{tabular}
\end{table}

\noindent These verdicts are not measurements; they are demonstrations that the rule is well-defined and that its output depends on the contract, not on the surface text.
That is exactly the property current pipelines lack: each case is one a regime-agnostic scorer must get wrong in at least one direction, because it has no slot for $O$, $\sigma$, or $\kappa$.

\section{Research Agenda (Open Problems)}\label{sec:agenda}

If hallucination is regime-relative, evaluation must make the regime a first-class object.
The framework and the audit above turn that principle into concrete, tractable problems, organized from measurement foundations (items~1--4) to system-level design (items~5--7).
We present these as the paper's main forward contribution: not a finished evaluator, but the specification one would have to build.

\textbf{Operationalizing \TR\ as a scorable contract.}
A \TR-aware benchmark must annotate $O$, $\sigma$, and $\kappa$ independently rather than folding them into a task label.
The graded components are not hypothetical to measure: stylistic latitude already has established datasets and metrics \citep{pavlick_empirical_2016, liu_semisupervised_2022, osyvokon_pavlick_formality_2016}, and $\sigma$ could be placed on a continuous scale via pairwise preference judgments fitted to a Bradley--Terry or Elo model, as arena-style leaderboards already do for overall quality \citep{boubdir_elo_nodate}.
The open challenge is reliability: regime boundaries involve normative choices (what \emph{should} be grounded?) and creativity judgments are evaluator-dependent \citep{acar_creativity_2019, marco_reader_2025}; where agreement is lowest (typically on the usefulness dimension of \PD), the disagreement is itself informative and should be reported through soft labels rather than forced into consensus.

\textbf{Operationalizing \PD\ as a scored quantity.}
\PD\ is the framework's central new construct, yet Section~\ref{sec:framework} specifies it without saying how to measure it.
The three conditions of the labeling rule are not equally hard, and separating them turns \PD\ from a definition into a scoring protocol.
The \emph{licensing} condition is mechanical: it is read off the $\kappa$ annotation, so under $\kappa{=}0$ no claim can be \PD\ and the score is zero by construction, while under $\kappa{>}0$ one checks only whether the claim falls inside the licensed scope (e.g., the declared fictional frame).
The \emph{presented-on-the-regime's-terms} condition is a stance check: whether a claim is marked as a hypothesis or asserted as verified fact is a verifiable, constraint-style property of the text, close to the format constraints that instruction-following evaluation already scores \citep{zhou_ifeval_2023}, and far more tractable than judging content quality.
The hard component is \emph{usefulness}, the standard novelty-with-value criterion of creativity assessment \citep{runco_standard_2012, acar_creativity_2019, diedrich_are_2015}: novelty admits automated proxies (semantic-distance and LLM-judge scoring of divergent output reach strong human correlation \citep{organisciak_beyond_2023}), and value can be aggregated from pairwise preference judgments onto a continuous scale via Bradley--Terry or Elo models, as quality leaderboards already do \citep{noauthor_bradleyterry_2025, boubdir_elo_nodate}.
A claim then scores as \PD\ when it passes the licensing gate and the stance check and its usefulness exceeds a regime-conditioned threshold.
Because usefulness is evaluator-dependent \citep{marco_reader_2025}, \PD\ should be reported as a graded, soft-labeled quantity rather than a hard count, with disagreement on the usefulness dimension recorded rather than averaged away.
This is a proposed instrument, not a validated one; establishing its reliability is itself part of the agenda.

\textbf{Inferring \TR\ from prompts, and separating it from instruction-following.}
Users rarely state a formal regime, so $\TR(p)$ must be inferred from prompt text \citep{wang_user-centric_2024, bodonhelyi_user_2024}, and calibration is critical: misreading a strict prompt as permissive suppresses warranted penalties, while the reverse penalizes legitimate \PD.
Inference must also be kept distinct from instruction-following fidelity: whether a model satisfied a verifiable format constraint is separately measurable \citep{zhou_ifeval_2023} and should not be confused with whether its content respects the regime, since an output can follow the requested form while violating $\kappa$.

\textbf{Metrics for \emph{excess} \Hallu\ under a given \TR.}
Under a regime-aware view what matters is not the raw hallucination count but the excess relative to the regime's tolerance, which requires severity weighting (a fabricated safety claim outweighs a date slip), claim- vs.\ output-level aggregation, and regime-conditioned baselines (the acceptable rate differs between grounded summarization and brainstorming).
Severity-aware and risk-weighted proposals exist \citep{bang_hallulens_2025, wei_measuring_2024}, but none yet conditions severity on an explicit \TR\ contract.

\textbf{Robust \Hallu\ detection under stylistic variation.}
This is where $\sigma$ does its work in the framework: it changes no verdict, but it is the axis along which a verifier's \Hallu\ count can be corrupted.
Because raising $\sigma$ alone can move what a verifier flags, NLI-based verifiers risk conflating figurative language with unsupported claims \citep{chen_menli_2023, aynetdinov_semscore_2024}, a brittleness already visible in style-transfer metric meta-evaluations \citep{lai_multidimensional_2023, pauli_mind_2025}.
A diagnostic suite pairing identical propositional content with varying stylistic treatments \citep{jin_deep_2022} would separate verifier sensitivity to form from genuine divergence in content.

\textbf{Controlling \Hallu\ without collapsing \PD.}
Mitigation (constrained decoding, retrieval augmentation, post-hoc verification) typically targets \Hallu\ reduction without measuring collateral damage to creative quality, and the trade-off is asymmetric: verification-style interventions can preserve divergent thinking while some decoding constraints suppress it \citep{banerjee_does_2025, sun_hallucinating_2024, he_shakespearean_2025}.
A \TR-aware protocol should therefore report \emph{both} \Hallu\ reduction and \PD\ preservation under the same regime.

\textbf{Benchmarks crossing task $\times\,\sigma\,\times\,\kappa$.}
No existing benchmark systematically crosses task type, stylistic latitude, and invention license as independent variables; \textsc{CreativityPrism} and LitBench provide partial coverage but condition no scores on an explicit \TR\ \citep{hou_creativityprism_2025, fein_litbench_2025}.
A concrete next step is a claim-level annotation layer with labels such as supported, hallucination, productive divergence, unverifiable, and stylistic variation, evaluated as a research object in its own right, including whether it changes model rankings or instead chiefly sharpens error diagnosis.

\section{Discussion}\label{sec:discussion}

\subsection{Implications for evaluation practice}
The practical consequence of \TR-relative evaluation is a shift from defect counting to \emph{regime-conditioned diagnosis}.
Rather than asking ``how much does the output hallucinate?'', evaluation should ask ``does its divergence exceed what the \TR\ licenses?''
Three verdicts follow: \emph{aligned} (divergence matches the regime), \textit{excessive-}\textbf{\Hallu} (unsupported claims where grounding is required), and \textit{insufficient-}\textbf{\PD} (output too conservative for a permissive regime).
The first two read off the claim-level labels (a claim either is or is not \Hallu); the third does not, because too little divergence is an \emph{absence} that no per-claim label can record.
\textit{Insufficient-}\PD\ is therefore an output-level judgment that compares the output's licensed-divergence yield against what a permissive regime leads one to expect, the comparison creativity benchmarks already make when they penalize conservative outputs \citep{hou_creativityprism_2025}; the claim-level rule alone cannot deliver it.
The same measured rate therefore carries different meaning under strict and permissive contracts, so regime-specific baselines are not a refinement but a precondition for interpreting any score.
Even under permissive regimes, evaluation should still flag \emph{truth-commitment violations}, cases where invented content is presented as fact rather than as hypothesis or creative proposal \citep{bang_hallulens_2025, hou_creativityprism_2025}.

\subsection{Limitations of the position}
Our argument is conceptual, and four limitations bound it.
First, the framework is partly \emph{normative}: deciding what $O$ ought to be, and what a regime \emph{should} ground, is a value judgment about a task, not a measurement we can read off the text, so reasonable annotators will disagree on regime boundaries.
Second, $\TR(p)$ is an idealization that ambiguous prompts underdetermine; locating the contract at the prompt is a modeling choice, and inferring it reliably is an open problem (Section~\ref{sec:agenda}), not a solved preprocessing step.
Third, \PD\ is specified and we sketch a scoring protocol for it (Section~\ref{sec:agenda}), but we do not validate that protocol: its usefulness condition inherits the evaluator dependence of creativity assessment, so what we offer is a labeling rule and a measurement proposal, not an instrument shown to measure \PD\ reliably at scale.
Fourth, we argue this position rather than establishing it with a controlled study; the audit and worked verdicts show that the gap is real and that supplying $(O,\sigma,\kappa)$ resolves verdicts, but quantifying how much regime annotation changes evaluation outcomes is left to future empirical work.

\section{Conclusion}\label{sec:conclusion}

Hallucination is not a binary property of text but a violation relative to a prompt's truth regime.
The vocabulary introduced here, $\TR(p)=(O,\sigma,\kappa)$ together with \PD\ and \Hallu, makes explicit what current hallucination and creativity evaluations leave implicit: which oracle governs truth, how much stylistic latitude is granted, and whether new truth-conditional content is licensed.
Auditing widely used benchmarks shows that none of them encodes this contract, and applying the labeling rule by hand shows that supplying it flips verdicts the same output would otherwise receive by accident of protocol.
The agenda is therefore to move evaluation from defect counting to regime-conditioned diagnosis, so that it can penalize harmful fabrication under strict regimes while preserving productive divergence where exploration is the point.
The remaining work is to build the contract into benchmarks and to measure how much it changes what we conclude about models, but the first step is to stop scoring divergence as if every prompt asked for the same thing.

\appendix

\section{Formal Claim-Level Decision Rule}\label{app:decision-rule}
\noindent\textbf{Input:} a prompt $p$, an output $y$, and an oracle $O$.\\
\noindent\textbf{Output:} a label for every atomic claim in $y$.

\begin{enumerate}
    \item \textbf{Specify the regime.} Infer or state $\TR(p)=(O,\sigma,\kappa)$: what is truth-conditional, what is evidence-bound, what is optionally speculative, and what (if anything) may be invented.
    \item \textbf{Decompose.} Split $y$ into atomic claims, ideally at claim-level granularity.
    \item \textbf{Label each claim $c$:}
    \begin{itemize}
        \item[\Hallu] if $c$ is truth-conditional under $\TR(p)$ but unsupported or incorrect relative to $O$;
        \item[\PD] else if $\TR(p)$ permits $c$ and it improves usefulness or appropriateness while remaining constraint-compliant;
        \item[--] else \emph{neither}: purely stylistic variation or benign noise.
    \end{itemize}
\end{enumerate}
\noindent The supporting evidence for each step is given where the rule is introduced in Section~\ref{sec:framework}.

\bibliography{references}

\end{document}
